# Sampling Distribution of the Least Squares Estimators and Testing the Slope

> "If you do not know how to ask the right question, you discover nothing." - W. E. Demining





## Properties of the constants $k_i$ and $c_i$

In the previous section, we examined how the least squares estimators are linear combinations of the response variable $y$. Let's know look at the properties of the coefficients in @eq-w1_5 and @eq-w1_6. We will not present the proofs in this course but they are not complicated. 


The coefficients $k_{i}$ have the following properties:
$$
\begin{align}
\sum k_{i} & =0
\end{align}
$${#eq-w1_7}

$$
\begin{align}
\sum k_{i}x_i & =1 
\end{align}
$${#eq-w1_8}


$$
\begin{align}
\sum k_{i}^{2} & =\frac{1}{\sum \left(x_i-\bar{x}\right)^{2}}
\end{align}
$${#eq-w1_9}

Likewise, the coefficients $c_{i}$ have the following properties:
$$
\begin{align}
\sum c_{i} & =1 
\end{align}
$${#eq-w1_10}

$$
\begin{align}
\sum c_{i}x_i & =0 
\end{align}
$${#eq-w1_11}

$$
\begin{align}
\sum c_{i}^{2} & =\frac{1}{n}+\frac{\left(\bar{x}\right)^{2}}{\sum \left(x_i-\bar{x}\right)^{2}} 
\end{align}
$${#eq-w1_12}



::: {.callout-note}
## Review: The Expected Value of a Linear Combination

Recall that the expected value of a linear combination of the random variable $Y$ is
$$
E(aY+b)=aE(Y)+b
$$
where $a$ and $b$ are constants.

:::


## Expected Values of $b_0$ and  $b_1$

Before finding the expectations, recall $E\left(y_i\right)=\beta_{0}+\beta_{1}x_i$.

### Expected Value of $b_1$

The expected value of $b_1$ is
$$
\begin{align*}
E\left[b_1\right] & =E\left[\underbrace{\sum k_{i}y_i}_{(3.1)}\right]\\
& =\sum k_{i}\left(\beta_{0}+\beta_{1}x_i\right)\\
& =\beta_{0}\underbrace{\sum k_{i}}_{(4.1)}+\beta_{1}\underbrace{\sum k_{i}x_i}_{(4.2)}\\
& =\beta_{1}
\end{align*}
$$

### Expected Value of $b_0$


The expected value of $b_0$ is
$$
\begin{align*}
E\left[b_0\right] & =E\left[\underbrace{\sum c_{i}y_i}_{(3.2)}\right]\\
& =\sum c_{i}\left(\beta_{0}+\beta_{1}x_i\right)\\
& =\beta_{0}\underbrace{\sum c_{i}}_{(4.4)}+\beta_{1}\underbrace{\sum c_{i}x_i}_{(4.5)}\\
& =\beta_{0}
\end{align*}
$$

Therefore, $b_0$ is an *unbiased* estimator of $\beta_0$ and $b_1$ is an *unbiased* estimator of $\beta_1$.

::: {.callout-note}
## Review: Unbiased Estimator

Recall that an unbiased estimator for some parameter is an estimator that has an expected value equal to that parameter.
:::




## Variances of $b_0$ and $b_1$

To find the variances, we will use a result from mathematical statistics:
Let $Y_{1},\ldots,Y_{n}$ be uncorrelated random variables and let
$a_{1},\ldots,a_{n}$ be constants. Then
$$
\begin{align}
Var\left[\sum a_{i}Y_i\right] & =\sum a_{i}^{2}Var\left[Y_i\right]
\end{align}
$${#eq-w1_13}


Recall that we assume the response variables $y_i$'s are *independent*.

Technically, we assume the $y_i$'s are *uncorrelated*. In general, uncorrelated does not imply independent. However, if the random variables are jointly normally distributed (recall our third assumption of the model), then uncorrelated does imply independent.


Also, note that
$$
\begin{align*}
    Var\left[Y\right]& = Var\left[\beta_0 + \beta_1 x + \varepsilon\right]\\
    & = Var\left[\varepsilon\right]\\
    & = \sigma^2
\end{align*}
$$

### Variance of $b_1$

The variance of $b_1$ is
$$
\begin{align}
Var\left[b_1\right] & =Var\left[\underbrace{\sum k_{i}y_i}_{(3.1)}\right]\\
& =\underbrace{\sum k_{i}^{2}}_{(4.3)}Var\left[y_i\right]\\
& =\frac{\sigma^{2}}{\sum \left(x_i-\bar{x}\right)^{2}}
\end{align}
$${#eq-w1_14}

### Variance of $b_0$

The variance of $b_0$ is
$$
\begin{align}
Var\left[b_0\right] & =Var\left[\underbrace{\sum c_{i}y_i}_{(3.2)}\right]\nonumber\\
& =\underbrace{\sum c_{i}^{2}}_{(4.6)}Var\left[y_i\right]\\
& =\sigma^{2}\left[\frac{1}{n}+\frac{\left(\bar{x}\right)^{2}}{\sum \left(x_i-\bar{x}\right)^{2}}\right]
\end{align}
$${#eq-w1_15}

## Best Linear Unbiased Estimators (BLUEs)



We see from @eq-w1_5 and @eq-w1_6 that $b_0$ and $b_1$
are *linear* estimators.



Any estimator for $\beta_{1}$, which we will
denote as $\hat{\beta}_{0}$, that takes the form
$$
\begin{align*}
\hat{\beta}_{1} & =\sum a_{i}y_i
\end{align*}
$$
where $a_{i}$ is some constant, is called a *linear estimator*.


Of all linear estimators for $\beta_0$ and $\beta_1$ that are *unbiased*, the least squares estimators, $b_0$ and $b_1$, have the smallest variance.

This is summarized in the following well known theorem:

::: {#thm-gauss}

## Gauss Markov Theorem

For the simple linear regression model, the least squares
estimators $b_0$ and $b_1$ are unbiased and have minimum
variance among all unbiased linear estimators.

:::







An estimator that is linear, unbiased, and has the smallest variance of all unbiased linear estimators is called the *best linear unbiased estimator* (BLUE).


::: {.callout-tip collapse="true"}
## For those who want to see the math:

Proof of the Gauss Markov Theorem:

For all linear estimators that are unbiased, we must have
$$
\begin{align*}
E\left[\hat{\beta}_{1}\right] & =E\left[\sum a_{i}y_i\right]\\
& =\sum a_{i}E\left[y_i\right]\\
& =\beta_{1}
\end{align*}
$$
Since $E\left[y_i\right]=\beta_{0}+\beta_{1}x_i$, then we must have
$$
\begin{align*}
E\left[\hat{\beta}_{1}\right] & =\sum a_{i}\left(\beta_{0}+\beta_{1}x_i\right)\\
& =\beta_{0}\sum a_{i}+\beta_{1}\sum a_{i}x_i\\
& =\beta_{1}
\end{align*}
$$
Therefore,
$$
\begin{align*}
\sum a_{i} & =0\\
\sum a_{i}x_i & =1
\end{align*}
$$
We now examine the variance of $\hat{\beta}_{1}$:
$$
\begin{align*}
Var\left[\hat{\beta}_{1}\right] & =\sum a_{i}^{2}Var\left[y_i\right]\\
& =\sigma^{2}\sum a_{i}^{2}
\end{align*}
$$
Let's now define $a_{i}=k_{i}+d_{i}$ where $k_{i}$ is defined in
@eq-w1_5.
and $d_{i}$ is some arbitrary constant.


We will show that adding
a constant (whether negative or positive) to $k_i$ cannot make the
variance smaller. Thus, the smallest variance of the linear estimator $\hat{\beta}_1$ is when $a_i=k_i$.

The variance of $\hat{\beta}_{1}$ can now be written as
$$
\begin{align*}
Var\left[\hat{\beta}_{1}\right] & =\sigma^{2}\sum a_{i}^{2}\\
& =\sigma^{2}\sum\left(k_{i}+d_{i}\right)^{2}\\
& =\sigma^{2}\sum\left(k_{i}^{2}+2k_{i}d_{i}+d_{i}^{2}\right)\\
& =Var\left[b_1\right]+2\sigma^{2}\sum k_{i}d_{i}+\sigma^{2}\sum d_{i}^{2}
\end{align*}
$$
Examining the second term and using the expression of $k_{i}$
in @eq-w1_5, we see that
$$
\begin{align*}
\sum k_{i}d_{i} & =\sum k_{i}\left(a_{i}-k_{i}\right)\\
& =\sum a_{i}k_{i}-\underbrace{\sum k_{i}^{2}}_{(4.3)}\\
& =\sum a_{i}\frac{x_i-\bar{x}}{\sum\left(x_i-\bar{x}\right)^{2}}-\frac{1}{\sum\left(x_i-\bar{x}\right)^{2}}\\
& =\frac{\sum a_{i}x_i-\bar{x}\sum a_{i}}{\sum\left(x_i-\bar{x}\right)^{2}}-\frac{1}{\sum\left(x_i-\bar{x}\right)^{2}}\\
& =\frac{1-\bar{x}\left(0\right)}{\sum\left(x_i-\bar{x}\right)^{2}}-\frac{1}{\sum\left(x_i-\bar{x}\right)^{2}}\\
& =0
\end{align*}
$$

We now have the variance of $\hat{\beta}_{1}$ as
$$
\begin{align*}
Var\left[\hat{\beta}_{1}\right] & =Var\left[b_1\right]+\sigma^{2}\sum d_{i}^{2}
\end{align*}
$$
This variance is minimized when $\sum d_{i}^{2}=0$ which only happens
when $d_{i}=0$.


Thus, the unbiased linear estimator with the smallest
variance is when $a_{i}=k_{i}$. That is, the least squares estimator
$b_1$ in @eq-w1_5 has the smallest variance of all unbiased
linear estimators of $\beta_{1}$.



A similar argument can be used to show that $b_0$ has the smallest variance of all unbiased
linear estimators of $\beta_{0}$.

:::





## Sampling Distribution for $b_1$


Now that we see that the least squares estimator $b_1$
is the BLUE for $\beta_{1}$, we will now examine the sampling distribution for $b_1$.


We previously discussed  that the mean of the sampling distribution of $b_1$
is
$$
E[b_1]=\beta_1
$$
with a variance of
$$
\begin{align}
Var\left[b_1\right]
& =\frac{\sigma^{2}}{\sum \left(x_i-\bar{x}\right)^{2}}
\end{align}
$${#eq-w1_14}



Note that in our model with our four assumptions, $y$ is normally distributed. That is,
$$
\begin{align}
    y\sim N\left(\beta_0+\beta_1 x, \sigma^2\right)
\end{align}
$${#eq-4_10}

To learn about the sampling distributions of the least squares estimators, we will use the following theorems from mathematical statistics:

::: {#thm-4_1}

## Sum of Independent Normal Random Variables

If
$$
Y_i\sim N\left(\mu_i,\sigma_i^2\right)
$$
are independent, then the linear combination $\sum_i a_iY_i$
is also normally distributed where $a_i$ are constants. In particular
$$
\sum_i a_iY_i \sim N\left(\sum_i a_i\mu_i, \sum_i a_i^2\sigma_i^2\right)
$$

:::


::: {#thm-4_2}

## Adding a Constant to a Normal Random Variable


If
$$
Y\sim N\left(\mu,\sigma^2\right)
$$
then for any real constant $c$,
$$
Y+c\sim N\left(\mu+c,\sigma^2\right)
$$

:::







Since $Y$
is normally distributed by @eq-4_10, then we can apply @thm-4_1  which
implies that $b_1$ is normally distributed. That is,
$$
\begin{align}
b_1 & \sim N\left(\beta_{1},\frac{\sigma^{2}}{\sum\left(x_i-\bar{x}\right)^{2}}\right)
\end{align}
$${#eq-w2_3}

### Standardized Score

Since $b_1$ is normally distributed, we can *standardize* it so that
the resulting statistic will have a standard normal distribution.


Therefore, we have
$$
\begin{align}
z=\frac{b_1-\beta_{1}}{\sqrt{\frac{\sigma^{2}}{\sum\left(x_i-\bar{x}\right)^{2}}}} & \sim N\left(0,1\right)
\end{align}
$${#eq-w2_4}


### Studentized Score

In practice, the standardized score $z$ is not useful since we do
not know the value of $\sigma^{2}$. We can estimate
$\sigma^{2}$ with the statistic
$$
s^2 = \frac{SSE}{n-2}
$$

Using this estimate for $\sigma^2$  leads us to a $t$-score:
$$
\begin{align}
t=\frac{b_1-\beta_{1}}{\sqrt{\frac{s^{2}}{\sum\left(x_i-\bar{x}\right)^{2}}}} & \sim t\left(n-2\right)
\end{align}
$${#eq-w2_4b}

We call this $t$ statistic, the *studentized* score.

::: {.callout-tip collapse="true"}
## For those who want to see the math:

It is important to note the following theorem from math stats presented
here without proof:

::: {#thm-04_3}

## Distribution of the sample variance of the residuals

For the sample variance of the residuals $s^{2}$, the quantity
\begin{align*}
\frac{\left(n-2\right)s^{2}}{\sigma^{2}} & =\frac{SSE}{\sigma^{2}}
\end{align*}
is distributed as a chi-square distribution with $n-2$ degrees of
freedom. That is,
\begin{align*}
\frac{SSE}{\sigma^{2}} & \sim\chi^{2}\left(n-2\right)
\end{align*}

:::


We will use another important theorem form math stats (again presented
without proof):

::: {#thm-04_4}

## Ratio of independent standard normal and chi-square statistics

If $Z\sim N\left(0,1\right)$ and $W\sim\chi^{2}\left(\nu\right)$,
and $Z$ and $W$ are independent, then the statistic
\begin{align*}
\frac{Z}{\sqrt{\frac{W}{\nu}}}
\end{align*}
is distributed as a Student's $t$ distribution with $\nu$ degrees
of freedom.

:::



We will take the standardized score in @eq-w2_4 and divide by
\begin{align*}
\sqrt{\frac{\frac{\left(n-2\right)s^{2}}{\sigma^{2}}}{\left(n-2\right)}} & =\sqrt{\frac{s^{2}}{\sigma^{2}}}
\end{align*}
to give us
\begin{align*}
t & =\frac{\frac{b_1-\beta_{1}}{\sqrt{\frac{\sigma^{2}}{\sum\left(x_i-\overline{X}\right)^{2}}}}}{\sqrt{\frac{s^{2}}{\sigma^{2}}}}\\
& =\frac{b_1-\beta_{1}}{\sqrt{\frac{\sigma^{2}}{\sum\left(x_i-\overline{X}\right)^{2}}}\sqrt{\frac{s^{2}}{\sigma^{2}}}}\\
& =\frac{b_1-\beta_{1}}{\sqrt{\frac{s^{2}}{\sum\left(x_i-\overline{X}\right)^{2}}}}
\end{align*}
which will have a Student's $t$ distribution with $n-2$ degrees
of freedom.

:::






## Assessing the Utility of the Model: Making Inferences About the Slope {#sec-modelutility}


Suppose
that the independent variable $x$ is completely unrelated to
the dependent variable $y$.

What could be said about the values of
$\beta_0$ and $\beta_1$ in the hypothesized probabilistic model
\begin{align*}
    y = \beta_0 +\beta_1 x + \varepsilon
\end{align*}
if $x$ contributes no information for the prediction of $y$?



The
implication is that the mean of $y$, does not change as $x$ changes.
 In other words, the line would just be a *horizontal* line.






If $E(y)$ does not change as $x$ increases, then using $x$ to predict $y$ in the linear model is not *useful*.



Regardless
of the value of $x$, you always predict the same value of $y$. In the
straight-line model, this means that the true slope, $\beta_1$, is equal to 0.

Therefore, to test the null hypothesis that $x$
contributes no information for the prediction of $y$ against the
alternative hypothesis that these variables are *linearly* related with a
slope differing from 0, we test
\begin{align*}
    H_0:\beta_1 = 0\\
    H_a:\beta_1\ne 0
\end{align*}



If the data support the alternative hypothesis, we conclude that $x$
does contribute information for the prediction of $y$ using the straight-line model (although the true relationship between $E(y)$ and $x$ could
be more complex than a straight line). Thus, to some extent, this is
a test of the *utility* of the hypothesized model.

The appropriate test statistic is the studentized score given above:
$$
\begin{align}
    t &= \frac{b_1-\beta_{1}}{\sqrt{\frac{s^{2}}{\sum\left(x_i-\overline{X}\right)^{2}}}}\\
    &=\frac{b_1}{\sqrt{\frac{s^{2}}{SS_{xx}}}}
\end{align}
$${#eq-4_14}




Another way to make inferences about the slope $\beta_1$ is to estimate it
using a *confidence* *interval*
$$
\begin{align}
    b_1 \pm \left(t_{\alpha/2}\right)s_{b_1}
\end{align}
$${#eq-4_15}
where
\begin{align*}
    s_{b_1} = \frac{s}{\sqrt{SS_{xx}}}
\end{align*}




We can obtain the p-value for the hypothesis test by using the `summary` function with an `lm` object. For the previous example consisting of the `mtcars` data.

::: {#exm-4_1}

## @exm-3_1 - revisited


```{r message=FALSE, warning=F}
library(tidyverse)

fit = lm(mpg~wt, data = mtcars)

```

We find the least squares estimates as

```{r message=FALSE, warning=F}
summary(fit)
```

From the output, we see the p-value is
$1.29\times 10^{-10}$. So we have sufficient evidence to conclude that the true population slope is different than zero.

To find the confidence interval, we can use the `confint` function with the `lm` object. 

```{r message=FALSE, warning=F}
confint(fit, level = 0.95)
```


We 95% confident that the true population slope is in the interval $`r paste0("(",round(confint(fit, level = 0.95)[2, 1], 3), ", ", round(confint(fit, level = 0.95)[2, 2],3), ")")`$

:::

