# Checking the Linearity and Constant Variance Assumptions {#sec-checking}

> "…the statistician knows…that in nature there never was a normal distribution, there never was a straight line, yet with normal and linear assumptions, known to be false, he can often derive results which match, to a useful approximation, those found in the real world." - George Box






## Residual Diagnostics


### Model Assumptions
 
 Let's review the assumptions for the simple linear regression model:

1. The *mean* of the probability distribution of $\varepsilon$ is 0.
2. The *variance* of the probability distribution of $\varepsilon$ is *constant* for all settings of the independent variable $x$. 
3. The probability distribution of $\varepsilon$ is *normal*.
4. The errors associated with any two different observations are *independent*.



After fitting the model, we will need to check these assumptions.

### Residuals
 
We check the assumptions of the model by examining the residuals:
$$
\begin{align}
{e_{i}  =y_{i}-\hat{y}_{i}}
\end{align}
$${#eq-w1_19}

We do this since the assumptions, with the exception of the linearity assumption,  are based on the error terms $\varepsilon_i$. We can think of $e_i$ as an observed value of $\varepsilon_i$.   


### Properties of Residuals

Below are some properties of the residuals:
$$
\begin{align}
\sum e_{i} & =0\\
\sum x_{i}e_{i} & =0\\
\sum\hat{y}_{i}e_{i} & =0 \\
\sum y_{i} & =\sum\hat{y}_{i}
\end{align}
$${#eq-w1_20-23}


Clearly, from @eq-w1_20-23, the mean of the residuals is
$$
\begin{align}
    \bar{e}_i=0
\end{align}
$${#eq-w3_1}


The variance of all $n$ residuals, $e_1,\ldots,e_n$ is
$$
\begin{align}
\frac{\sum\left(e_{i}-\bar{e}\right)^{2}}{n-2} & =\frac{\sum e_{i}^{2}}{n-2}\\
& =\frac{SSE}{n-2}\\
& =MSE\\
& =s^{2}
\end{align}
$${#eq-w3_2}






### Semistudentized Residuals

It will be helpful to *studentize* each residuals. As always, we do
this by subtracting off the mean, $\bar{e}_{i}$, and dividing by
the standard error of $e_{i}$.


We know by @eq-w3_1 that $\bar{e}_{i}=0$.


In  @eq-w3_2, we said the variance of the sample of the $e_{i}$'s is MSE.
For each individual $e_{i}$, the standard error is not quite $\sqrt{MSE}$.
The actual standard error is dependent on the predictor variable(s).
We will discuss this more in multiple regression.



For now, we will use the approximation
$\sqrt{MSE}$ and calculate
\begin{align*}
e_{i}^{*} & =\frac{e_{i}-\bar{e}}{\sqrt{MSE}}\\
& =\frac{e_{i}}{\sqrt{MSE}}
\end{align*}
We call $e_{i}^{*}$ the *semistudentized* residual since the standard
error is an approximation.










## The Linearity Assumption


### Residual Plots

We can check the linearity assumption by plotting the residuals vs the *predictor*   variable or plotting the residuals vs the  *fitted*  values.


We usually examine a *scatterplot* to determine if a linear relationship between $x$ and $y$ is appropriate. There are times when the scatterplot makes it difficult to see if a nonlinear relationship exists. This may be the case if the observed $y$ are close to the fitted line $\hat{y}_i$. This usually means the slope is steep.


::: {#exm-7_1}

## Weight and Height Data

In this example, we will consider the weights (in kg) and heights (in m) of 16 women ages 30-39. The dataset is from kaggle.


```{r message=FALSE, warning=F}
library(tidyverse)

dat = read_csv("Weight_Height.csv")

ggplot(dat, aes(x=Weight, y=Height))+
  geom_point()


#fit the model
fit = lm(Weight~Height, data=dat)

#plot with regression line
ggplot(dat, aes(x=Weight, y=Height))+
  geom_point()+
  geom_smooth(method="lm", formula=y~x, se=F)

#make dataset with Weight, the fitted values, and residuals
dat2 = tibble(x = dat$Weight, 
              yhat = fit$fitted.values, 
              e = fit$residuals)

#plot x by residuals
ggplot(dat2, aes(x=x, y=e))+
  geom_point()+
  geom_hline(yintercept = 0, col="red")

#plot fitted values by residuals
ggplot(dat2, aes(x=yhat, y=e))+
  geom_point()+
  geom_hline(yintercept = 0, col="red")

```

:::



### Plotting against Predictor Variable or Fitted Values

Plotting the residuals against $x$  will provide the same information as plotting the residuals against $\hat{y}$ for the simple linear regression model.


When more predictor variables are considered, then plotting against the $x$ variables and plotting against $\hat{y}$ may provide different information. It is usually helpful to plot both in that case.


### Data Transformation for Linearity

When the linearity assumption does not hold (as seen in the residual plots), then a *nonlinear* model may be considered or a *transformation* on either $x$ or $y$ can be attempted to make the relationship linear.


### Transforming $x$

Transforming the response variable $y$ may lead to issues with other assumptions such as the
*constant variance* assumption or the *normality* of $\varepsilon$ assumption.



If our only concern is the linearity assumption, then transforming $x$ will be the best option. This transformation may be a square root transformation $\sqrt{X}$, a log transformation $\log{X}$, or some power transformation $X^{p}$ were $p$ is some real number.


Sometimes a transformation of $x$ will not be enough to satisfy the linearity assumption. In that case, the simple linear regression model should be abandoned in favor of a nonlinear model.



## Homogeneity of Variance



### Residual Plots

As we did previously, we can plot the residuals against the predictor variable $x$ or against the fitted values $\hat{y}$ to help determine whether the variance of the error term $\varepsilon$ is *constant*.



When the variance is constant, we say the model has *homoscedasticity*. When the variance is nonconstant, we say the model has *heteroscedasticity*.



### Absolute Residuals and Squared Residuals

When examining a residual plot for non-constant variance, we look for any clear changes in the spread of the residuals. One common clear pattern seen when heteroscedasticity is present is a *cone* pattern.


We are usually not concerned about the sign of the residual when examining for heteroscedasticity. Thus, it is common to plot the *absolute* residuals or the *squared* residuals vs the predictor variable or fitted values.


Usually a least squares line is then fit to the absolute residual or squared residuals plot. If this line has a significant slope, then this gives evidence that the variance is nonconstant.


::: {#exm-w3_3_1}

## Diastolic Blood Pressure Data

We will model the diastolic blood pressure by the age of 54 healthy adult women. The data are found in Kutner et al[^1^].

[^1^]: Kutner, M. H., Nachtsheim, C. J., Neter, J., & Li, W. (2004). Applied Linear Statistical Models McGraw-Hill/lrwin series operations and decision sciences.

```{r message=FALSE, warning=F}
library(tidyverse)

dat = read.table("bloodpressure.txt", header=T)

fit = lm(dbp~age, data=dat)

#add the fit line without using geom_smooth
ggplot(dat, aes(x=age, y=dbp))+
  geom_point()+
  geom_abline(slope = fit$coefficients[2],
              intercept = fit$coefficients[1])

```


By examining the scatterplot of dbp vs age, we already see evidence of nonconstant variance.

```{r message=FALSE, warning=FALSE}
dat$e = fit |> resid()
dat$yhat = fit |> fitted()

#use the geom_smooth function to add a least squares line
#for the residuals, the least squares line will always be #horizontal at 0
ggplot(dat, aes(x=age, y=e))+
  geom_point()+
  geom_smooth(method = "lm",se = F)

#plot the squared residuals and add least squares line
ggplot(dat, aes(x=age, y=e^2))+
  geom_point()+
  geom_smooth(method = "lm",se = F)

#plot the squared residuals and add least squares line
ggplot(dat, aes(x=age, y=abs(e)))+
  geom_point()+
  geom_smooth(method = "lm",se = F)


```

In any of the residual plots we examine, we see the "cone" shape of the residuals which indicates the variance is nonconstant.

:::



### Tests for Heteroscedasticity

We can set up a hypothesis test for heteroscedasticity:
\begin{align*}
H_0:&\text{ the variance is constant}\\
H_a:&\text{ the variance is non-constant}
\end{align*}

The procedures we will use usually test for variance that increases or decreases over the values of $x$. That is, the spread of the points about the line is a cone shape.




### Levene's Test and Brown-Forsythe Test

The *Levene test*[^2^] starts by dividing the range of the predictor variable $x$ into $k$ intervals. For each of the intervals, calculate the mean of the residuals in that interval $\bar{e}_{j}$ where $j$ denotes the $j$th interval.

[^2^]: Levene, H. (1960) Robust Tests for Equality of Variances. In: Olkin, I., Ed., *Contributions to Probability and Statistics*, Stanford University Press, Palo Alto, 278-292.


Now, define the absolute deviation from the mean
$$
d_{ik}=|e_{ik}-\bar{e}_{j}|
$$

An ANOVA F-test is then performed on the $k$ groups. The ANOVA F-test will be discussed more later in the course. A small p-value is evidence that the variance is non-constant over the values of $x$.


The *Brown-Forsythe test*[^3^] is a modification of the Levene test in which the median $\tilde{e}_k$ is used instead of the mean $\bar{e}_l$. This test is robust against nonnormal errors.

[^3^]: Brown, M. B., & Forsythe, A. B. (1974). Robust tests for the equality of variances. *Journal of the American statistical association*, 69(346), 364-367.



### Breusch-Pagan Test

A second test for non-constant variance is the *Breusch-Pagan* test[^4^]. This test assumes the variance for each $\varepsilon_i$ is related to the values of $x$ in the following way:
$$
\ln \sigma^2_i = \gamma_0 + \gamma_1 X_i
$$
Hence, a linear regression is assumed between $\sigma^2$ and $x$. We can fit the line by squaring the residuals and regressing on $x$. The   third plot in example above shows the squared residuals plotted against $x$ and the fitted line.

[^4^]: Breusch, T. S., & Pagan, A. R. (1979). A simple test for heteroscedasticity and random coefficient variation. *Econometrica: Journal of the econometric society*, 1287-1294.


If the variance is non-constant, then the slope of this line will be non-zero. Thus, a test for the slope is conducted. A small p-value is evidence that the variance is non-constant.


Because it is testing the slope, the Brown-Forsythe test assumes the error terms are independent and normally distributed.



::: {#exm-w3_3_2}

## @exm-w3_3_1 revisited 


Let's examine the blood pressure data from @exm-w3_3_1 again.

If you want the Brown-Forsythe test, then you will need to determine the $k$ groups yourself and then use the `bf.test` function in `onewaytests` package.



In the `lmtest` package, the Breusch-Pagan test can be conducted by using the `bptest` function.

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(lmtest)

dat = read.table("bloodpressure.txt", header=T)

fit = lm(dbp~age, data=dat)

#Breusch-Pagan test
bptest(dbp~age, data=dat)

```


Since the p-value is low, then there is sufficient evidence to conclude the variance is not constant through different values of $x$.

:::



