# The Regression Model in Matrix Terms

> "The essence of mathematics is not to make simple things complicated, but to make complicated things simple." - S. Gudder







## The Matrices for the Different Components


Recall the regression model is
\begin{align*}
y_{i}= & \beta_{0}+\beta_{1}x_{i1}+\beta_{2}x_{i2}+\cdots+\beta_{p-1}x_{i,p-1}+\varepsilon_{i}\\
& \varepsilon\overset{iid}{\sim}N\left(0,\sigma^{2}\right)
\end{align*}
for $i=1,\ldots,n$.




This implies:
\begin{align*}
y_{1} & =\beta_{0}+\beta_{1}x_{11}+\beta_{2}x_{12}+\cdots+\beta_{p-1}x_{1,p-1}+\varepsilon_{1}\\
y_{2} & =\beta_{0}+\beta_{1}x_{21}+\beta_{2}x_{22}+\cdots+\beta_{p-1}x_{2,p-1}+\varepsilon_{2}\\
& \vdots\\
y_{n} & =\beta_{0}+\beta_{1}x_{n1}+\beta_{2}x_{n2}+\cdots+\beta_{p-1}x_{n,p-1}+\varepsilon_{n}
\end{align*}





### Response Vector

We define the response vector as 
\begin{align*}
\underset{{n\times1}}{\textbf{Y}}=\left[\begin{array}{c}
y_{1}\\
y_{2}\\
\vdots\\
y_{n}
\end{array}\right]
\end{align*}




### Random Error Vector

We define the vector of random errors as 
\begin{align*}
\underset{{n\times1}}{\boldsymbol{\varepsilon}}=\left[\begin{array}{c}
\varepsilon_{1}\\
\varepsilon_{2}\\
\vdots\\
\varepsilon_{n}
\end{array}\right]
\end{align*}





### Vector of Coefficients

We define the vector of coefficients as
\begin{align*}
\underset{{p\times1}}{\boldsymbol{\beta}}=\left[\begin{array}{c}
\beta_{0}\\
\beta_{1}\\
\vdots\\
\beta_{p-1}
\end{array}\right]
\end{align*}




### The Design Matrix

We define the matrix of the predictor variables as 
\begin{align*}
\underset{{n\times p}}{\textbf{X}}=\left[\begin{array}{ccccc}
1 & x_{11} & x_{12} & \cdots & x_{1,p-1}\\
1 & x_{21} & x_{22} & \cdots & x_{2,p-1}\\
\vdots & \vdots & \vdots & \ddots & \vdots\\
1 & x_{n1} & x_{n2} & \cdots & x_{n,p-1}
\end{array}\right]
\end{align*}

Note that the first column of $\bf{X}$ is a vector of ones. This column will represent the coefficient of the $y$-intercept in the model. 

::: {.example title="Example 11.1: Building the response vector and design matrix in R"}

In chapter 10, we fit the multiple regression model

$$
Volume_i=\beta_0+\beta_1Girth_i+\beta_2Height_i+\varepsilon_i
$$

using the `trees` dataset. In matrix notation, this model uses a response vector ${\bf Y}$ and a design matrix ${\bf X}$.

```{r}
trees_matrix_fit <- lm(Volume ~ Girth + Height, data = trees)

Y_trees <- as.matrix(trees["Volume"])
X_trees <- model.matrix(trees_matrix_fit)
b_trees <- as.matrix(coef(trees_matrix_fit))

matrix_dimensions <- tibble::tibble(
  object = c("Y_trees", "X_trees", "b_trees"),
  role = c("response vector", "design matrix", "estimated coefficient vector"),
  dimension = c(
    paste(dim(Y_trees), collapse = " x "),
    paste(dim(X_trees), collapse = " x "),
    paste(dim(b_trees), collapse = " x ")
  )
)

knitr::kable(matrix_dimensions)
```

The design matrix begins with an intercept column of ones.

```{r}
head(X_trees) |>
  knitr::kable(digits = 3)
```

The column of ones is what allows the matrix multiplication ${\bf X}{\boldsymbol\beta}$ to include the intercept term $\beta_0$.

:::








## The Model

We can now write the model as

\begin{align*}
\underset{{n\times1}}{\textbf{Y}} & =\underset{{n\times p}}{\textbf{X}}\underset{{p\times 1}}{\boldsymbol{\beta}}+\underset{{n\times1}}{\boldsymbol{\varepsilon}}
\end{align*}
since:
\begin{align*}
\left[\begin{array}{c}
y_{1}\\
y_{2}\\
\vdots\\
y_{n}
\end{array}\right] & =\left[\begin{array}{ccccc}
1 & x_{11} & x_{12} & \cdots & x_{1,p-1}\\
1 & x_{21} & x_{22} & \cdots & x_{2,p-1}\\
\vdots & \vdots & \vdots & \ddots & \vdots\\
1 & x_{n1} & x_{n2} & \cdots & x_{n,p-1}
\end{array}\right]\left[\begin{array}{c}
\beta_{0}\\
\beta_{1}\\
\vdots\\
\beta_{p-1}
\end{array}\right]+\left[\begin{array}{c}
\varepsilon_{1}\\
\varepsilon_{2}\\
\vdots\\
\varepsilon_{n}
\end{array}\right]\\
& =\left[\begin{array}{c}
\beta_{0}+\beta_{1}x_{11}+\beta_2x_{12}+\cdots+\beta_{p-1}x_{1,p-1}\\
\beta_{0}+\beta_{1}x_{21}+\beta_2x_{22}+\cdots+\beta_{p-1}x_{2,p-1}\\
\vdots\\
\beta_{0}+\beta_{1}x_{n1}+\beta_2x_{n2}+\cdots+\beta_{p-1}x_{n,p-1}
\end{array}\right]+\left[\begin{array}{c}
\varepsilon_{1}\\
\varepsilon_{2}\\
\vdots\\
\varepsilon_{n}
\end{array}\right]\\
& =\left[\begin{array}{c}
\beta_{0}+\beta_{1}x_{11}+\beta_2x_{12}+\cdots+\beta_{p-1}x_{1,p-1}+\varepsilon_{1}\\
\beta_{0}+\beta_{1}x_{21}+\beta_2x_{22}+\cdots+\beta_{p-1}x_{2,p-1}+\varepsilon_{2}\\
\vdots\\
\beta_{0}+\beta_{1}x_{n1}+\beta_2x_{n2}+\cdots+\beta_{p-1}x_{n,p-1}+\varepsilon_{n}
\end{array}\right]
\end{align*}

::: {.callout-note collapse="false" title="The dimension check tells the story"}

The matrix model

$$
{\bf Y}={\bf X}{\boldsymbol\beta}+{\boldsymbol\varepsilon}
$$

has compatible dimensions:

$$
\underset{n\times1}{\bf Y}
=
\underset{n\times p}{\bf X}
\underset{p\times1}{\boldsymbol\beta}
+
\underset{n\times1}{\boldsymbol\varepsilon}.
$$

The product ${\bf X}{\boldsymbol\beta}$ is an $n\times1$ vector, so it has one model mean for each observation.

:::




### Multivariate Normal Distribution

The assumption on the normal error model for the random error term
is
\begin{align*}
\varepsilon\overset{iid}{\sim} & N\left(0,\sigma^{2}\right).
\end{align*}
In matrix notation, this can be expressed with the \blank{multivariate} normal
distribution. 




Note that the univariate normal distribution has a probability density
function expressed as
\begin{align*}
f\left(x\right) & =\frac{1}{\sigma\sqrt{2\pi}}\exp\left[-\frac{1}{2\sigma^{2}}\left(x-\mu\right)^{2}\right]
\end{align*}
where $\mu$ is the mean of the distribution and $\sigma$ is the
standard deviation.




The multivariate normal distribution is expressed as
\begin{align*}
f\left({\bf Y}\right) & =\frac{1}{\left(2\pi\right)^{n/2}\left|\boldsymbol{\Sigma}\right|^{1/2}}\exp\left[-\frac{1}{2}\left({\bf Y}-\boldsymbol{\mu}\right)^{\prime}\boldsymbol{\Sigma}^{-1}\left({\bf Y}-\boldsymbol{\mu}\right)\right]
\end{align*}
where ${\bf Y}$ is a $n\times1$ vector, $\boldsymbol{\mu}$ is a
$n\times1$ vector of means, and $\boldsymbol{\Sigma}$ is the $n\times n$
covariance matrix. 





We denote the multivariate normal distribution of a random vector
${\bf Y}$ as
\begin{align*}
{\bf Y} & \sim N_{n}\left(\boldsymbol{\mu},\boldsymbol{\Sigma}\right).
\end{align*}

For the normal error model, the mean vector of the random vector $\boldsymbol{\varepsilon}$
is a vector of zeros (${\bf 0}$) and the covariance matrix is
\begin{align*}
{\bf Cov}\left(\boldsymbol{\varepsilon}\right) & =\left[\begin{array}{cccc}
\sigma^{2} & 0 & \cdots & 0\\
0 & \sigma^{2} & \cdots & 0\\
\vdots & \vdots & \ddots & \vdots\\
0 & 0 & \cdots & \sigma^{2}
\end{array}\right]\\
& =\sigma^{2}{\bf I}.
\end{align*}

::: {.callout-note collapse="false" title="Scalar and matrix assumptions side-by-side"}

The matrix notation is not adding new regression assumptions. It is rewriting the familiar scalar assumptions in a compact form.

| Scalar statement | Matrix statement | Meaning |
|---|---|---|
| $E(\varepsilon_i)=0$ | $E({\boldsymbol\varepsilon})={\bf 0}$ | The errors average to 0. |
| $Var(\varepsilon_i)=\sigma^2$ | Diagonal entries of $Cov({\boldsymbol\varepsilon})$ are $\sigma^2$ | Each error has the same variance. |
| $Cov(\varepsilon_i,\varepsilon_j)=0$ for $i\ne j$ | Off-diagonal entries of $Cov({\boldsymbol\varepsilon})$ are 0 | Different errors are uncorrelated. |
| $\varepsilon_i$ are normally distributed | ${\boldsymbol\varepsilon}\sim N_n({\bf 0},\sigma^2{\bf I})$ | The whole error vector follows a multivariate normal distribution. |

Because ${\bf Y}={\bf X}{\boldsymbol\beta}+{\boldsymbol\varepsilon}$ and ${\bf X}$ is treated as fixed, these assumptions imply

$$
E({\bf Y})={\bf X}{\boldsymbol\beta}
\qquad\text{and}\qquad
Cov({\bf Y})=\sigma^2{\bf I}.
$$

:::





### The Model in Matrix Notation

We now represent the *normal* *errors* multiple regression model as
$$
\begin{align}
{\textbf{Y}}= & {\textbf{X}}{\boldsymbol{\beta}}+{\boldsymbol{\varepsilon}}\\
& \boldsymbol{\varepsilon} \sim N_{n}\left({\bf 0},\sigma^{2}{\bf I}\right)
\end{align} 
$${#eq-w4_18}

Note that 
\begin{align*}
\textbf{E}\left(\textbf{Y}\right) & =\textbf{X}\boldsymbol{\beta}
\end{align*}

::: {.example title="Example 11.2: The fitted mean vector"}

In practice, we estimate ${\boldsymbol\beta}$ with ${\bf b}$. Then ${\bf X}{\bf b}$ gives the fitted values.

```{r}
fitted_by_matrix <- X_trees %*% b_trees

fitted_mean_compare <- tibble::tibble(
  observation = 1:6,
  matrix_fitted = as.numeric(fitted_by_matrix[1:6, 1]),
  lm_fitted = as.numeric(fitted(trees_matrix_fit)[1:6]),
  observed_volume = trees$Volume[1:6]
)

knitr::kable(fitted_mean_compare, digits = 4)
```

The matrix product ${\bf X}{\bf b}$ gives the same fitted values as `lm()`. The observed responses do not usually fall exactly on these fitted values because the residuals are not usually zero.

:::



## Least Squares and Inferences Using Matrices







### Least Squares Estimators

For the model in @eq-w4_18
 we want to minimize the squared distances between the observed $y_i$ and the fitted $\hat{y}_i$. 




We will express the sum of the squared distances in @eq-w4_3
 in matrix terms as
 $$
\begin{align}
Q & =\left({\bf Y}-{\bf X}{\bf b}\right)^{\prime}\left({\bf Y}-{\bf X}{\bf b}\right)
\end{align}
$${#eq-w4_19}
where ${\bf Y}$ is the response vector, ${\bf X}$ is the design
matrix, and ${\bf b}$ is the vector of estimators
\begin{align*}
{\bf b} & =\left[\begin{array}{c}
b_{0}\\
b_{1}\\
\vdots\\
b_{p-1}
\end{array}\right]
\end{align*}





### Minimizing the Sums of Squares

We will minimize $Q$ in @eq-w4_19 by taking the derivative with respect
to ${\bf b}$ and setting it equal to the vector of zeros. The derivative
is done using the properties listed in @sec-multiple and using the
chain rule:
\begin{align*}
\frac{\partial\left({\bf Y}-{\bf X}{\bf b}\right)^{\prime}\left({\bf Y}-{\bf X}{\bf b}\right)}{\partial{\bf b}} & =-2{\bf X}^{\prime}\left({\bf Y}-{\bf X}{\bf b}\right)
\end{align*}

Setting this derivative equal to the vector of zeros gives us
\begin{align*}
-2{\bf X}^{\prime}\left({\bf Y}-{\bf X}{\bf b}\right)={\bf 0} & \Longrightarrow{\bf X}^{\prime}\left({\bf Y}-{\bf X}{\bf b}\right)={\bf 0}
\end{align*}





### The Normal Equations

We can distribute ${\bf X}^{\prime}$ through the parentheses
$$
\begin{align}
{\bf X}^{\prime}\left({\bf Y}-{\bf X}{\bf b}\right)={\bf 0} & \Longrightarrow{\bf X}^{\prime}{\bf Y}-{\bf X}^{\prime}{\bf X}{\bf b}={\bf 0}\\
& \Longrightarrow{\bf X}^{\prime}{\bf X}{\bf b}={\bf X}^{\prime}{\bf Y}
\end{align}
$${#eq-w4_20}
Equation @eq-w4_20 represents the *normal equations* in matrix format.




### The Estimators

Solving the normal equations for ${\bf b}$ gives us the \blank{least squares}
estimators:
$$
\begin{align}
{\bf X}^{\prime}{\bf X}{\bf b}={\bf X}^{\prime}{\bf Y} & \Longrightarrow\left({\bf X}^{\prime}{\bf X}\right)^{-1}{\bf X}^{\prime}{\bf X}{\bf b}=\left({\bf X}^{\prime}{\bf X}\right)^{-1}{\bf X}^{\prime}{\bf Y}\\
& \Longrightarrow{\bf b}=\left({\bf X}^{\prime}{\bf X}\right)^{-1}{\bf X}^{\prime}{\bf Y}
\end{align}
$${#eq-w4_21}

::: {.example title="Example 11.3: Computing the coefficient vector with matrices"}

For the `trees` model, the coefficient vector can be calculated directly from

$$
{\bf b}=({\bf X}'{\bf X})^{-1}{\bf X}'{\bf Y}.
$$

```{r}
b_by_matrix <- solve(t(X_trees) %*% X_trees) %*% t(X_trees) %*% Y_trees

coefficient_comparison <- tibble::tibble(
  term = rownames(b_by_matrix),
  matrix_formula = as.numeric(b_by_matrix),
  lm_estimate = as.numeric(coef(trees_matrix_fit)),
  difference = as.numeric(b_by_matrix) - as.numeric(coef(trees_matrix_fit))
)

knitr::kable(coefficient_comparison, digits = 10)
```

The matrix formula and `lm()` give the same coefficient estimates. The matrix form is not a different method; it is a compact way to write ordinary least squares for any number of predictors.

:::







::: {.example title="Example 11.4: A singular design matrix from a redundant predictor"}

The formula

$$
{\bf b}=({\bf X}'{\bf X})^{-1}{\bf X}'{\bf Y}
$$

requires ${\bf X}'{\bf X}$ to have an inverse. If one predictor is a perfect copy or exact multiple of another predictor, the columns of ${\bf X}$ are redundant and ${\bf X}'{\bf X}$ is singular.

For example, create a new predictor that is exactly twice `Girth`.

```{r}
trees_redundant <- transform(trees, Girth_copy = 2 * Girth)

X_singular <- model.matrix(Volume ~ Girth + Girth_copy + Height, data = trees_redundant)
XtX_singular <- t(X_singular) %*% X_singular

inverse_message <- tryCatch(
  {
    solve(XtX_singular)
    "Inverse exists"
  },
  error = function(e) "No inverse: X'X is singular"
)

singular_summary <- tibble::tibble(
  quantity = c("columns in X", "rank of X", "full column rank?", "inverse check"),
  result = c(
    as.character(ncol(X_singular)),
    as.character(qr(X_singular)$rank),
    as.character(qr(X_singular)$rank == ncol(X_singular)),
    inverse_message
  )
)

knitr::kable(singular_summary)
```

The design matrix has more columns than its rank because `Girth_copy` contains no new information beyond `Girth`. This is why the inverse in the least-squares formula does not exist.

```{r}
singular_fit <- lm(Volume ~ Girth + Girth_copy + Height, data = trees_redundant)

broom::tidy(singular_fit) |>
  knitr::kable(digits = 4)
```

R can still detect the redundancy and fit the estimable part of the model, but the matrix inverse formula fails because the coefficient vector is not uniquely determined.

:::

### Fitted Values

The fitted values are
$$
\begin{align}
\hat{{\bf Y}} & ={\bf X}{\bf b}
\end{align}
$${#eq-w4_22}

Substituting @eq-w4_21 for ${\bf b}$ gives us
$$
\begin{align}
\hat{{\bf Y}}={\bf X}\left({\bf X}^{\prime}{\bf X}\right)^{-1}{\bf X}^{\prime}{\bf Y}
\end{align}
$${#eq-w4_23}

The $n\times n$ matrix ${\bf X}\left({\bf X}^{\prime}{\bf X}\right)^{-1}{\bf X}^{\prime}$
pops up a number of times in multiple regression analysis. We call
this matrix the *hat matrix* and denote it as
$$
\begin{align}
{\bf H} & ={\bf X}\left({\bf X}^{\prime}{\bf X}\right)^{-1}{\bf X}^{\prime}
\end{align}
$${#eq-w4_24}

Thus, the fitted values can be expressed as
$$
\begin{align}
\hat{{\bf Y}} & ={\bf H}{\bf Y}
\end{align}
$${#eq-w4_25}

::: {.example title="Example 11.5: The hat matrix maps observed responses to fitted values"}

The hat matrix is called the "hat" matrix because it puts the hat on ${\bf Y}$:

$$
\hat{\bf Y}={\bf H}{\bf Y}.
$$

```{r}
H_trees <- X_trees %*% solve(t(X_trees) %*% X_trees) %*% t(X_trees)
fitted_by_hat <- H_trees %*% Y_trees

hat_comparison <- tibble::tibble(
  observation = 1:6,
  hat_matrix_fitted = as.numeric(fitted_by_hat[1:6, 1]),
  lm_fitted = as.numeric(fitted(trees_matrix_fit)[1:6]),
  difference = as.numeric(fitted_by_hat[1:6, 1]) - as.numeric(fitted(trees_matrix_fit)[1:6])
)

knitr::kable(hat_comparison, digits = 10)
```

The hat matrix is an $n\times n$ matrix. For the `trees` data, it is `r nrow(H_trees)` by `r ncol(H_trees)`, because it transforms the whole response vector into the whole fitted-value vector.

:::

::: {.callout-note collapse="false" title="The diagonal of the hat matrix"}

The diagonal entries of ${\bf H}$ are called *leverages*. Leverage measures how unusual an observation's predictor values are compared with the rest of the data.

This chapter focuses on the matrix form of fitted values. Later, leverage will become important when we study influential observations.

:::




::: {.example title="Example 11.6: Fitted values as a projection"}

Geometrically, least squares projects the observed response vector ${\bf Y}$ onto the column space of ${\bf X}$. The fitted vector $\hat{\bf Y}$ is the point in that column space closest to ${\bf Y}$, and the residual vector ${\bf e}={\bf Y}-\hat{\bf Y}$ is perpendicular to that space.

The picture below is a two-dimensional cartoon of the idea. In a real regression problem, the vectors may live in a much higher-dimensional observation space.

```{r}
#| fig-width: 6
#| fig-height: 4

plot(
  NA,
  xlim = c(-0.3, 4),
  ylim = c(-0.4, 2.8),
  xlab = "",
  ylab = "",
  axes = FALSE,
  asp = 1,
  main = "Least squares as projection"
)

abline(h = 0, col = "lightblue", lwd = 8)
text(3.45, -0.25, "Column space of X", col = "steelblue")

arrows(0, 0, 3, 0, length = 0.1, lwd = 2, col = "steelblue")
arrows(0, 0, 3, 2, length = 0.1, lwd = 2, col = "darkgreen")
arrows(3, 0, 3, 2, length = 0.1, lwd = 2, col = "firebrick")

points(c(3, 3), c(0, 2), pch = 19, col = c("steelblue", "darkgreen"))

text(3.2, 0.12, expression(hat(bold(Y))), col = "steelblue")
text(2.35, 2.15, expression(bold(Y)), col = "darkgreen")
text(3.25, 1, expression(bold(e) == bold(Y) - hat(bold(Y))), col = "firebrick")
text(1.3, 0.25, expression(hat(bold(Y)) == bold(H) * bold(Y)), col = "steelblue")

segments(2.85, 0, 2.85, 0.15, col = "gray40")
segments(2.85, 0.15, 3, 0.15, col = "gray40")
```

This visual explains why the normal equations give ${\bf X}'{\bf e}={\bf 0}$: after projection, the leftover residual vector is orthogonal to every direction contained in the column space of ${\bf X}$.

:::

### Residuals

The residuals can be expressed in matrix terms as
$$
\begin{align}
{\bf e} & ={\bf Y}-\hat{{\bf Y}}\\
& ={\bf Y}-{\bf X}{\bf b}
\end{align}
$${#eq-w4_26}

Using @eq-w4_25, the residuals can be expressed in terms of the hat matrix
as well
$$
\begin{align}
{\bf e} & ={\bf Y}-{\bf H}{\bf Y}\\
& =\left({\bf I}-{\bf H}\right){\bf Y}
\end{align}
$${#eq-w4_27}

::: {.example title="Example 11.7: Least-squares residuals are orthogonal to the design matrix"}

The normal equations imply

$$
{\bf X}'{\bf e}={\bf 0}.
$$

This means the least-squares residual vector is orthogonal to every column of the design matrix, including the intercept column.

```{r}
e_trees <- Y_trees - fitted_by_hat
orthogonality_check <- t(X_trees) %*% e_trees

orthogonality_table <- tibble::tibble(
  column_of_X = rownames(orthogonality_check),
  X_prime_e = as.numeric(orthogonality_check)
)

knitr::kable(orthogonality_table, digits = 10)
```

The values are essentially zero, up to tiny numerical rounding error. This is one of the most important geometric facts about least squares: the residual vector is perpendicular to the space spanned by the predictor columns.

:::




### Estimator for the Variance

We have seen in simple regression
 that $s^{2}$ is an estimate of the variance
$\sigma^{2}$. Recall that we call $s^{2}$ the MSE which is just
the SSE divided by the degrees of freedom ($n-2$ in simple linear
regression). 



In matrix terms, we can express SSE as
$$
\begin{align}
SSE & ={\bf e}^{\prime}{\bf e}\\
& ={\bf Y}^{\prime}\left({\bf I}-{\bf H}\right)^{\prime}\left({\bf I}-{\bf H}\right){\bf Y}\\
& ={\bf Y}^{\prime}\left({\bf I}-{\bf H}\right){\bf Y}
\end{align}
$${#eq-w4_28}

The term $\left({\bf I}-{\bf H}\right)^{\prime}\left({\bf I}-{\bf H}\right)$ simplifies to $\left({\bf I}-{\bf H}\right)$ since $\left({\bf I}-{\bf H}\right)$ is idempotent[^1]


[^1]: Idempotent matrix is a square matrix which when multiplied by itself, gives back the same matrix.




The MSE is
\begin{align*}
MSE & =\frac{SSE}{n-p}
\end{align*}

Here the degrees of freedom are now $n-p$ since there are $p$ parameters
$\left(\beta_{0},\beta_{1},\ldots,\beta_{p-1}\right)$ that need to
be estimated first in order to get $SSE$. As before, MSE is an estimator
for $\sigma^{2}$.

::: {.example title="Example 11.8: Computing SSE and MSE with matrices"}

Because ${\bf e}$ is the residual vector, the error sum of squares can be written as

$$
SSE={\bf e}'{\bf e}.
$$

```{r}
sse_matrix <- as.numeric(t(e_trees) %*% e_trees)
n_trees <- nrow(X_trees)
p_trees <- ncol(X_trees)
mse_matrix <- sse_matrix / (n_trees - p_trees)

mse_comparison <- tibble::tibble(
  quantity = c("SSE from e'e", "MSE from SSE / (n - p)", "MSE from lm summary"),
  value = c(
    sse_matrix,
    mse_matrix,
    summary(trees_matrix_fit)$sigma^2
  )
)

knitr::kable(mse_comparison, digits = 6)
```

For this model, $n=`r n_trees`$ and $p=`r p_trees`$, so the error degrees of freedom are $n-p=`r n_trees - p_trees`$.

:::

## Recap

This chapter rewrote the multiple regression model using matrix notation.

| Idea | Meaning |
|---|---|
| **Response vector** | ${\bf Y}$ is the $n\times1$ vector of observed responses. |
| **Error vector** | ${\boldsymbol\varepsilon}$ is the $n\times1$ vector of random errors. |
| **Coefficient vector** | ${\boldsymbol\beta}$ is the $p\times1$ vector of regression parameters. |
| **Design matrix** | ${\bf X}$ is the $n\times p$ matrix containing the intercept column and predictor columns. |
| **Matrix regression model** | ${\bf Y}={\bf X}{\boldsymbol\beta}+{\boldsymbol\varepsilon}$. |
| **Normal error model** | ${\boldsymbol\varepsilon}\sim N_n({\bf 0},\sigma^2{\bf I})$. |
| **Matrix assumptions** | The scalar error assumptions can be summarized by $E({\boldsymbol\varepsilon})={\bf 0}$ and $Cov({\boldsymbol\varepsilon})=\sigma^2{\bf I}$. |
| **Least-squares criterion** | $Q=({\bf Y}-{\bf X}{\bf b})'({\bf Y}-{\bf X}{\bf b})$. |
| **Normal equations** | ${\bf X}'{\bf X}{\bf b}={\bf X}'{\bf Y}$. |
| **Least-squares estimator** | ${\bf b}=({\bf X}'{\bf X})^{-1}{\bf X}'{\bf Y}$, when the inverse exists. |
| **Singular design** | If a predictor is an exact linear combination of other predictors, ${\bf X}'{\bf X}$ is singular and the inverse does not exist. |
| **Fitted values** | $\hat{\bf Y}={\bf X}{\bf b}$. |
| **Hat matrix** | ${\bf H}={\bf X}({\bf X}'{\bf X})^{-1}{\bf X}'$, so $\hat{\bf Y}={\bf H}{\bf Y}$. |
| **Projection view** | Least squares projects ${\bf Y}$ onto the column space of ${\bf X}$ to produce $\hat{\bf Y}$. |
| **Residual vector** | ${\bf e}={\bf Y}-\hat{\bf Y}=({\bf I}-{\bf H}){\bf Y}$. |
| **MSE** | $MSE=SSE/(n-p)$ estimates $\sigma^2$. |

## Check your understanding

::: {.callout-note collapse="false" title="Problems"}

1. What are the dimensions of ${\bf Y}$ in the matrix regression model?

2. Why does the design matrix ${\bf X}$ usually include a first column of ones?

3. If a model has an intercept and two predictors, how many columns does ${\bf X}$ have?

4. In the matrix model ${\bf Y}={\bf X}{\boldsymbol\beta}+{\boldsymbol\varepsilon}$, what does ${\bf X}{\boldsymbol\beta}$ represent?

5. What does the assumption ${\boldsymbol\varepsilon}\sim N_n({\bf 0},\sigma^2{\bf I})$ say about the errors?

6. How do the scalar assumptions $E(\varepsilon_i)=0$, $Var(\varepsilon_i)=\sigma^2$, and $Cov(\varepsilon_i,\varepsilon_j)=0$ appear in matrix notation?

7. Why is $Q=({\bf Y}-{\bf X}{\bf b})'({\bf Y}-{\bf X}{\bf b})$ the matrix version of the sum of squared residuals?

8. What are the normal equations in matrix form?

9. Why must ${\bf X}'{\bf X}$ be invertible for the formula ${\bf b}=({\bf X}'{\bf X})^{-1}{\bf X}'{\bf Y}$ to work?

10. Give an example of how a redundant predictor can make ${\bf X}'{\bf X}$ singular.

11. What is the hat matrix, and why is it called the hat matrix?

12. What does it mean to say that least squares projects ${\bf Y}$ onto the column space of ${\bf X}$?

13. What does the residual vector ${\bf e}$ measure?

14. What does ${\bf X}'{\bf e}={\bf 0}$ mean conceptually?

15. Why is the denominator of MSE equal to $n-p$ in multiple regression?

16. What information is contained on the diagonal of the hat matrix?

17. How does matrix notation help when moving from simple regression to multiple regression?

:::

::: {.callout-tip collapse="true" title="Solutions"}

1. **$n\times1$.** The response vector has one row for each observation and one column.

2. **To include the intercept.** The first column of ones is multiplied by $\beta_0$, which adds the same intercept term to every observation's model mean.

3. **Three columns.** There is one intercept column plus one column for each of the two predictors, so $p=3$.

4. **The model mean vector.** The product ${\bf X}{\boldsymbol\beta}$ contains the expected response values implied by the regression model.

5. **Zero mean, common variance, no covariance.** The errors have mean 0, common variance $\sigma^2$, and covariance matrix $\sigma^2{\bf I}$, which represents no covariance between different errors.

6. **They become vector and matrix statements.** The assumptions become $E({\boldsymbol\varepsilon})={\bf 0}$ and $Cov({\boldsymbol\varepsilon})=\sigma^2{\bf I}$. The zero vector gives the mean-zero assumption; the diagonal $\sigma^2$ entries give common variance; the off-diagonal zeros give zero covariance between different errors.

7. **It multiplies residuals by themselves and sums.** The vector ${\bf Y}-{\bf X}{\bf b}$ contains residuals. Multiplying its transpose by itself gives the sum of squared residuals.

8. **${\bf X}'{\bf X}{\bf b}={\bf X}'{\bf Y}$.** These equations come from setting the derivative of the least-squares criterion equal to zero.

9. **The inverse is needed.** The formula solves the normal equations by multiplying by $({\bf X}'{\bf X})^{-1}$. If that inverse does not exist, the formula cannot produce a unique coefficient vector.

10. **Exact redundancy causes singularity.** If one predictor is exactly twice another predictor, or if a copied predictor is included twice, then one column of ${\bf X}$ is an exact linear combination of another. The model cannot uniquely separate those effects.

11. **It maps observed values to fitted values.** The hat matrix is ${\bf H}={\bf X}({\bf X}'{\bf X})^{-1}{\bf X}'$. It is called the hat matrix because $\hat{\bf Y}={\bf H}{\bf Y}$.

12. **Closest fitted vector.** The column space of ${\bf X}$ contains all possible fitted vectors ${\bf X}{\bf b}$. Least squares chooses the fitted vector in that space closest to the observed response vector ${\bf Y}$.

13. **Observed minus fitted.** The residual vector contains the differences between observed responses and fitted values.

14. **Residuals are orthogonal to the predictor columns.** The residual vector has no remaining linear association with any column of the design matrix under the least-squares fit.

15. **Because $p$ parameters were estimated.** The model uses $p$ degrees of freedom to estimate the coefficients, leaving $n-p$ error degrees of freedom.

16. **Leverage.** The diagonal entries of the hat matrix measure leverage, or how unusual each observation's predictor values are.

17. **It scales the notation.** Matrix notation lets us write one formula for any number of predictors instead of writing separate equations for every coefficient.

:::













