---
editor: 
  markdown: 
    wrap: 72
---

# Model Assumptions and the ANOVA F-test

> "Regression analysis is the hydrogen bomb of the statistics
> arsenal." - Charles Wheelan

The steps we followed in developing a simple linear model are applicable
to the multiple regression model.

**Analyzing a Multiple Regression Model**

1.  Collect the *sample data* (i.e., the value of $y$, $x_1$, $x_2$,
    ..., $x_k$) for each unit in the sample.
2.  Hypothesize the *form* of the model (i.e., the deterministic
    component), $E(y)$. This involves choosing which independent
    variables to include in the model.
3.  Use the method of least squares to *estimate* the unknown
    *parameters*.
4.  Specify the probability distribution of the random error component
    $\varepsilon$ and estimate its *variance* $\sigma^2$.
5.  Statistically evaluate the *utility* of the model.
6.  Check that the *assumptions* on $\varepsilon$ are satisfied and make
    model modifications, if necessary.
7.  Finally, if the model is deemed adequate, use the fitted model to
    estimate the *mean* value of $y$ or to *predict* a particular value
    of $y$ for given values of the independent variables, and to make
    other inferences.

We will first assume the model is known and the assumptions hold. We
will then come back to step 2 and step 6 later. Let's first discuss what
the assumptions are.

## Model Assumptions

The assumptions for the multiple regression model are the same as the
simple linear model. That is,

-   *Linearity* - We assume the model is linear in the *parameters* but
    not necessarily linear in the *predictor* variables. Thus, some of
    the predictor variables may need to be transformed.
-   *constant* *variance*
-   *normality*
-   *independence*

The inferences on the multiple regression model will depend on these
assumptions holding. We will discuss how to check these later.

::: {.callout-note collapse="false" title="Assumptions and the F-test"}

The ANOVA $F$-test is a model-utility test, but it still relies on the regression assumptions.

- Linearity affects whether the fitted model is describing the right mean structure.
- Constant variance affects the reliability of the standard error estimate.
- Normality is especially important for small-sample inference.
- Independence affects whether the information in the data is being counted correctly.

A small p-value from the global $F$-test does not prove that the model assumptions are satisfied. It only says that, under the model assumptions, the predictors collectively explain more variation than we would expect by chance if all slope coefficients were 0.

:::

## A First-Order Model with Quantitative Predictors

Recall that a *first-order* model means that none of the predictor
variables are functions of any other predictor variables.

When the independent variables are quantitative, the $\beta$ parameters
in the first-order model have similar interpretations as the simple
regression model. The difference is that when we interpret the $\beta$
that multiplies one of the variables (e.g., $x_1$), we must be certain
to hold the values of the remaining independent variables (e.g., $x_2$,
$x_3$) *fixed*.

::: {.callout-warning collapse="false" title="Partial slopes are still conditional"}

The ANOVA $F$-test asks whether the predictors are collectively useful. It does not change how individual slopes are interpreted.

In a first-order model with quantitative predictors, each slope is interpreted as the expected change in the response for a one-unit increase in that predictor, *holding the other predictors fixed*.

:::

## Testing the Utility of a Model: The Analysis of Variance $F$-Test

The objective of step 5 in a multiple regression analysis is to conduct
a test of the *utility* of the model—that is, a test to determine
whether the model is adequate for predicting $y$.

Later, we will examine how to conduct $t$-tests on each $\beta$
parameter in a model, where $$
\begin{align*}
    H_0: \beta_j = 0, \qquad j=1, 2, \ldots, p-1
\end{align*}
$$

However, this approach is generally not a good way to determine whether
the *overall* model is contributing information for the prediction of
$y$. If we were to conduct a series of $t$-tests to determine whether
the independent variables are contributing to the predictive
relationship, we would be very likely to make one or more errors in
deciding which terms to retain in the model and which to exclude.

Suppose you fit a first-order model with 10 quantitative independent
variables, $x_1, x_2,..., x_{10}$, and decide to conduct $t$-tests on
all 10 individual $\beta$’s in the model, each at $\alpha = .05$.

For any one test, the probability of making a type I error is 0.05: $$
\begin{align*}
    P(\text{Reject } H_0|\beta_j=0) &= {0.05}\\
    & {= 1 - 0.95}
\end{align*}
$$ If we were to do ten of these tests (one for each predictor
variable), then the probability that at least one is a type I error is
$$
\begin{align*}
    P(\text{Reject at least one } H_0|\beta_1=\beta_2=\cdots=\beta_{p-1}=0)
    & = 1-[(1-\alpha)^{10}]\\
    & {= 1-(0.95)^{10}}\\
    &{ = 0.401}
\end{align*}
$$

Even if all the $\beta$ parameters (except $\beta_0$) in the model are
equal to 0, approximately 40% of the time you will incorrectly reject
the null hypothesis at least once and conclude that some $\beta$
parameter is nonzero. In other words, the overall Type I error is about
.40, not .05.

To illustrate this inflated Type I error, let's look at the following
example.

::: {#exm-12_1 .example title="Inflated Type I error from many individual tests"}
We will simulate a sample of size 200 with a response variable $y$ and
ten predictor variables $x_1, x_2, \ldots, x_{10}$. The random error
term $\varepsilon$ will be a standard normal random variable.

The true model will have each coefficient $\beta$ set to 0, with the
exception of $\beta_0$ which will be 20. We will conduct a $t$-test for
each coefficient (except $\beta_0$). Since the true coefficient is 0, we
expect to see p-values greater than 0.05 for each coefficient.

We will fit the model and test the coefficients 1000 times. We will
count the number of times at least one of the coefficients was less than
0.05 (which would lead to a Type I error).

```{r warning=FALSE, message=FALSE}
library(tidyverse)

set.seed(3430)

n = 200

type1error = numeric(1000)

for(i in 1:1000){
  
  x = runif(n*10) |> matrix(ncol = 10)
  
  eps = rnorm(n)
  
  y = 20 + 0*x[,1] + 0*x[,2] + 0*x[,3] + 0*x[,4] +
    0*x[,5] + 0*x[,6] + 0*x[,7] + 0*x[,8] + 0*x[,9] +
    0*x[,10] + eps
  
  dat = tibble(y, x1 = x[,1], x2 = x[,2], x3 = x[,3],
               x4 = x[,4], x5 = x[,5], x6 = x[,6], x7 = x[,7],
               x8 = x[,8], x9 = x[,9], x10 = x[,10],)
  
  fit = lm(y~., data = dat)
  
  type1error[i] = any(summary(fit)$coefficients[-1,4] < 0.05)
}

mean(type1error)

```

We see that 41.4% of the time, we had at least one coefficient have a
p-value smaller than 0.05. This corresponds to the probability of at
least one Type I error of around 0.401.
:::

::: {.example title="Worked example: Comparing many t-tests to the global F-test"}

The previous simulation counted how often at least one individual $t$-test was significant when all slope coefficients were truly 0. Now compare that behavior with the global $F$-test.

```{r warning=FALSE, message=FALSE}
set.seed(3386)

nsim <- 1000
n <- 200
alpha <- 0.05

simulation_results <- tibble(
  many_t_tests = logical(nsim),
  global_f_test = logical(nsim)
)

for (i in 1:nsim) {
  x <- matrix(runif(n * 10), ncol = 10)
  eps <- rnorm(n)
  y <- 20 + eps

  dat <- tibble(
    y,
    x1 = x[, 1],
    x2 = x[, 2],
    x3 = x[, 3],
    x4 = x[, 4],
    x5 = x[, 5],
    x6 = x[, 6],
    x7 = x[, 7],
    x8 = x[, 8],
    x9 = x[, 9],
    x10 = x[, 10]
  )

  fit <- lm(y ~ ., data = dat)

  coefficient_p_values <- summary(fit)$coefficients[-1, 4]
  f_statistic <- summary(fit)$fstatistic
  global_p_value <- pf(
    f_statistic["value"],
    f_statistic["numdf"],
    f_statistic["dendf"],
    lower.tail = FALSE
  )

  simulation_results$many_t_tests[i] <- any(coefficient_p_values < alpha)
  simulation_results$global_f_test[i] <- global_p_value < alpha
}

type1_comparison <- simulation_results |>
  summarize(
    `At least one individual t-test is significant` = mean(many_t_tests),
    `Global F-test is significant` = mean(global_f_test)
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "method",
    values_to = "type_i_error_rate"
  )

knitr::kable(type1_comparison, digits = 3)
```

When all slope coefficients are truly 0, the global $F$-test keeps the Type I error rate close to $\alpha = 0.05$. The "many separate $t$-tests" approach has a much larger chance of producing at least one false positive.

:::

In multiple regression models for which a large number of independent
variables are being considered, conducting a series of $t$-tests may
cause the experimenter to include a large number of insignificant
variables and exclude some useful ones. If we want to test the utility
of a multiple regression model, we will need a *global* *test* (one that
encompasses all the $\beta$ parameters).

::: {.callout-note collapse="false" title="What the global F-test asks"}

For a model with predictors $x_1,\ldots,x_{p-1}$, the global $F$-test asks:

$$
H_0:\beta_1=\beta_2=\cdots=\beta_{p-1}=0
$$

against

$$
H_a:\text{at least one slope coefficient is not 0.}
$$

The test is global because it evaluates the model's predictors as a group. If the test is significant, it does not tell us which predictor is responsible. It only tells us that the model with predictors improves on the intercept-only model.

:::

### Partitioning the Sum of Squares

Recall when we discussed the coefficient of determination that we used
$SS_{yy}$ to denote the variability of the response variable $y$ from
its mean $\bar y$ (without regard to the model involving $x$).

Another name for $SS_{yy}$ is the *sum of squares total* (SSTO). We call
it this since it gives us a measure of total variability in $y$.

In multiple regression, SSTO is still the same as $SS_{yy}$. In matrix
notation this can be expressed as $$
\begin{align}
SSTO & ={\bf Y}^{\prime}{\bf Y}-\left(\frac{1}{n}\right){\bf Y}^{\prime}{\bf J}{\bf Y}
\end{align}
$$ {#eq-w4-34}

#### Sum of Squares Error (SSE)

Also recall the variability of $y$ about the regression line (in simple
linear regression) was expressed by SSE. We can think of this as the
variability of $y$ remaining after explaining some of the variability
with the regression model.

In multiple regression, SSE is still the sum of the squared distances
between the response $y$ and the fitted model $\hat{y}$. Now, the fitted
model is the fitted *hyperplane* instead of a line.

The SSE can be expressed in matrix terms as $$
\begin{align}
SSE & =\left({\bf Y}-{\bf X}{\bf b}\right)^{\prime}\left({\bf Y}-{\bf X}{\bf b}\right)\\
& ={\bf Y}^{\prime}{\bf Y}-{\bf b}^{\prime}{\bf X}^{\prime}{\bf Y}
\end{align}
$$ {#eq-w3_35}

#### Sum of Squares Regression (SSR)

If SSTO is the total variability of $y$ (without regard to the
predictor variables), and SSE is the variability of $y$ left over after
explaining the variability of $y$ with the model (including the
predictor variables), we might want to know the variability of $y$
explained by the model.

We call the variability explained by the regression model the *sum of
squares regression* (SSR).

We will show below that SSR can be expressed as $$
\begin{align}
SSR & =\sum\left(\hat{y}_{i}-\bar{y}\right)^{2}
\end{align}
$$ {#eq-w4_36} which can be expressed in matrix terms as $$
\begin{align}
SSR & ={\bf b}^{\prime}{\bf X}^{\prime}{\bf Y}-\left(\frac{1}{n}\right){\bf Y}^{\prime}{\bf J}{\bf Y}
\end{align}
$$ {#eq-w4_37}

#### Components of SSTO

To see how SSTO, SSR, and SSE relate to each other, consider how SSTO is
a sum of squares of $y$ from its mean $\bar{y}$: $$
\begin{align*}
y_{i}-\bar{y}
\end{align*}
$$

We can add and subtract the fitted value $\hat{y}_{i}$ to get $$
\begin{align*}
y_{i}-\bar{y} & =y_{i}-\hat{y}_{i}+\hat{y}_{i}-\bar{y}\\
& =\left(y_{i}-\hat{y}_{i}\right)+\left(\hat{y}_{i}-\bar{y}\right)
\end{align*}
$$

Squaring both sides gives us $$
\begin{align*}
\left(y_{i}-\bar{y}\right)^{2} & =\left[\left(y_{i}-\hat{y}_{i}\right)+\left(\hat{y}_{i}-\bar{y}\right)\right]^{2}\\
& =\left(y_{i}-\hat{y}_{i}\right)^{2}+\left(\hat{y}_{i}-\bar{y}\right)^{2}+2\left(y_{i}-\hat{y}_{i}\right)\left(\hat{y}_{i}-\bar{y}\right)
\end{align*}
$$

Summing both sides gives us $$
\begin{align*}
\sum\left(y_{i}-\bar{y}\right)^{2} & =\sum\left(y_{i}-\hat{y}_{i}\right)^{2}+\sum\left(\hat{y}_{i}-\bar{y}\right)^{2}+2\sum\left(y_{i}-\hat{y}_{i}\right)\left(\hat{y}_{i}-\bar{y}\right)\\
& =\sum\left(y_{i}-\hat{y}_{i}\right)^{2}+\sum\left(\hat{y}_{i}-\bar{y}\right)^{2}+2\sum\hat{y}_{i}e_{i}-2\bar{y}\sum e_{i}
\end{align*}
$$

Note that $\sum\hat{y}_{i}e_{i}=0$ 
<!-- from @eq-w1_22  -->
and $\sum e_{i}=0$.
<!-- from @eq-w1_20. -->

Therefore, we have $$
\begin{align*}
\sum\left(y_{i}-\bar{y}\right)^{2} & =\sum\left(\hat{y}_{i}-\bar{y}\right)^{2}+\sum\left(y_{i}-\hat{y}_{i}\right)^{2}\\
SSTO & =SSR+SSE 
\end{align*}
$$ {#eq-4_38}

We call this the *decomposition* of SSTO.

::: {.example title="Worked example: Decomposing SSTO for the `trees` model"}

For the `trees` model with `Volume` predicted by `Girth` and `Height`, the total variation in `Volume` can be split into variation explained by the model and variation left in the residuals.

```{r warning=FALSE, message=FALSE}
trees_anova_fit <- lm(Volume ~ Girth + Height, data = trees)

trees_y <- trees$Volume
trees_yhat <- fitted(trees_anova_fit)
trees_resid <- resid(trees_anova_fit)

ssto_trees <- sum((trees_y - mean(trees_y))^2)
ssr_trees <- sum((trees_yhat - mean(trees_y))^2)
sse_trees <- sum(trees_resid^2)

ss_decomposition <- tibble(
  source = c("Total variation", "Explained by regression", "Unexplained error"),
  symbol = c("SSTO", "SSR", "SSE"),
  value = c(ssto_trees, ssr_trees, sse_trees)
)

knitr::kable(ss_decomposition, digits = 4)
```

The decomposition says

$$
SSTO = SSR + SSE.
$$

For this model, `r round(ssto_trees, 4)` is approximately equal to `r round(ssr_trees, 4)` + `r round(sse_trees, 4)`.

:::

::: {.example title="Worked example: Visualizing `SSTO = SSR + SSE`"}

The decomposition can also be shown visually. The total bar shows SSTO. The second bar breaks that same total into SSR and SSE.

```{r warning=FALSE, message=FALSE}
ss_visual <- tibble(
  bar = c("SSTO", "SSR + SSE", "SSR + SSE"),
  component = c("SSTO", "SSR", "SSE"),
  value = c(ssto_trees, ssr_trees, sse_trees)
)

ggplot(ss_visual, aes(x = bar, y = value, fill = component)) +
  geom_col(width = 0.6, color = "white") +
  scale_fill_manual(
    values = c(
      "SSTO" = "gray70",
      "SSR" = "steelblue",
      "SSE" = "orange"
    )
  ) +
  labs(
    x = NULL,
    y = "Sum of squares",
    fill = "Component",
    title = "Total variation decomposes into explained and unexplained variation",
    subtitle = "For least squares regression with an intercept: SSTO = SSR + SSE"
  ) +
  theme_minimal()
```

The visual point is simple: the total variation in the response can be split into the part explained by the fitted model and the part left in the residuals.

:::

#### Degrees of Freedom

The degrees of freedom can be decomposed as well. Note that the degrees
of freedom for SSTO is $$
\begin{align*}
df_{SSTO} & =n-1
\end{align*}
$$ since the mean of $y$ is needed to be estimated with $\bar{y}$.

The degrees of freedom for SSE is $$
\begin{align*}
df_{SSE} & =n-p
\end{align*}
$$ since the $p$ coefficients $\beta_{0},\ldots,\beta_{p-1}$ need to be
estimated with $\hat{\beta}_{0},\ldots,\hat{\beta}_{p-1}.$

For SSR, the degrees of freedom is $$
\begin{align*}
df_{SSR} & =p-1
\end{align*}
$$ since there are $p$ estimated coefficients
$\hat{\beta}_{0},\ldots,\hat{\beta}_{p-1}$ but need to estimate the mean
of $y$ with $\bar{y}$.

Decomposing the degrees of freedom give us $$
\begin{align}
n-1 & =p-1+n-p\\
df_{SSTO} & =df_{SSR}+df_{SSE}
\end{align}
$$ {#eq-w4_29}

### The Analysis of Variance (ANOVA) Table

The sums of squares and degrees of freedom are commonly displayed in an
analysis of variance (ANOVA) table:

|   Source   |     df      |   SS   | MS  |  F  | p-value |
|:----------:|:-----------:|:------:|:---:|:---:|:-------:|
| Regression | $df_{SSR}$  | $SSR$  |     |     |         |
|   Error    | $df_{SSE}$  | $SSE$  |     |     |         |
|   Total    | $df_{SSTO}$ | $SSTO$ |     |     |         |



#### Mean Squares

Recall that if we divide SSE by its degrees of freedom, we obtain the
mean square error: 
$$
\begin{align}
MSE & =\frac{SSE}{n-p}
\end{align}
$${#eq-w4_40}


Likewise, if we divide SSR by its degrees of freedom, we obtain the
*mean square regression*: 
$$
\begin{align}
MSR & =\frac{SSR}{p-1}
\end{align} 
$${#eq-w4_41}

These values are also included in the ANOVA table:

|   Source   |     df      |   SS   | MS  |  F  | p-value |
|:----------:|:-----------:|:------:|:---:|:---:|:-------:|
| Regression | $df_{SSR}$  | $SSR$  |  $MSR$   |     |         |
|   Error    | $df_{SSE}$  | $SSE$  | $MSE$    |     |         |
|   Total    | $df_{SSTO}$ | $SSTO$ |     |     |         |


Note that although the sum of squares and degrees of freedom decompose,
the mean squares do not. That is 
$$
\begin{align*}
\frac{SSTO}{n-1} & \ne MSR+MSE
\end{align*}
$$

In fact, the mean square for the total ($SSTO/n-1$) does
not usually show up on the ANOVA table.

### The ANOVA F-test

We tested the slope in simple regression to see if there is a
significant linear relationship between $X$ and $Y$.

In multiple regression, we will want to see if there is any significant
linear relationship between any of the $x$s and $y$. Thus, we want to
test the hypotheses 
$$
\begin{align*}
H_{0}: & \beta_{1}=\beta_{2}=\cdots=\beta_{p-1}=0\\
H_{a}: & \text{at least one } \beta \text{ is not equal to zero}
\end{align*}
$$

To construct a test statistic, we first note that 
$$
\begin{align*}
\frac{SSE}{\sigma^{2}} & \sim\chi^{2}\left(n-p\right)
\end{align*}
$$
Also, if $H_{0}$ is true, then 
$$
\begin{align*}
\frac{SSR}{\sigma^{2}} & \sim\chi^{2}\left(p-1\right)
\end{align*}
$$

The ratio of two independent chi-square random variables divided by
their degrees of freedom gives a statistic that follows an F-distribution.

Since $SSE/\sigma^{2}$ and $SSR/\sigma^{2}$ are independent (proof not
given here), then under $H_{0}$, we can construct a test statistic as
$$
\begin{align}
F^{*} & =\left(\frac{\frac{SSR}{\sigma^{2}}}{p-1}\right)\div\left(\frac{\frac{SSE}{\sigma^{2}}}{n-p}\right)\\
& =\left(\frac{SSR}{p-1}\right)\div\left(\frac{SSE}{n-p}\right)\\
& =\frac{MSR}{MSE}
\end{align}
$${#eq-w4_42}


Large values of $F^{*}$ indicate evidence for $H_{a}$.




The test statistic and p-value are the last two components of the ANOVA
table:

|   Source   |     df      |   SS   | MS  |  F  | p-value |
|:----------:|:-----------:|:------:|:---:|:---:|:-------:|
| Regression | $df_{SSR}$  | $SSR$  |  $MSR$   | $F^{*}$    |  $P\left(F\ge F^{*}\right)$       |
|   Error    | $df_{SSE}$  | $SSE$  | $MSE$    |     |         |
|   Total    | $df_{SSTO}$ | $SSTO$ |     |     |         |

::: {.example title="Worked example: Building the ANOVA table by hand"}

Using the sums of squares from the `trees` model, we can construct the ANOVA table directly.

```{r warning=FALSE, message=FALSE}
n_trees_anova <- nobs(trees_anova_fit)
p_trees_anova <- length(coef(trees_anova_fit))

df_regression <- p_trees_anova - 1
df_error <- n_trees_anova - p_trees_anova
df_total <- n_trees_anova - 1

msr_trees <- ssr_trees / df_regression
mse_trees <- sse_trees / df_error
f_trees <- msr_trees / mse_trees
p_value_trees <- pf(f_trees, df_regression, df_error, lower.tail = FALSE)

manual_anova_table <- tibble(
  source = c("Regression", "Error", "Total"),
  df = c(df_regression, df_error, df_total),
  SS = c(ssr_trees, sse_trees, ssto_trees),
  MS = c(msr_trees, mse_trees, NA_real_),
  F = c(f_trees, NA_real_, NA_real_),
  p_value = c(p_value_trees, NA_real_, NA_real_)
)

knitr::kable(manual_anova_table, digits = 4)
```

The global $F$ statistic compares explained variation per regression degree of freedom to unexplained variation per error degree of freedom.

$$
F^*=\frac{MSR}{MSE}.
$$

Large values of $F^*$ occur when the regression model explains a lot of variation relative to the residual noise.

:::

::: {.callout-warning collapse="false" title="What a significant F-test does not tell you"}

A significant global $F$-test tells us that at least one slope coefficient is not 0. It does not tell us:

- which predictor or predictors are useful,
- whether the model assumptions are satisfied,
- whether the relationship is causal,
- whether the model predicts well on new data.

Those questions require follow-up coefficient tests, diagnostic plots, subject-area reasoning, and validation.

:::





::: {#exm-12_2 .example title="The global F-test for the `trees` model"}
We will fit a multiple regression model to the `trees` dataset. The Tidymodels framework will be used.

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(tidymodels)

data(trees)

# Prepare data
dat_recipe <- recipe(Volume ~ Girth + Height, data = trees)

# Set up model
lm_model <- linear_reg() |>
  set_engine("lm")

# Set up the workflow
lm_workflow <- workflow() |>
  add_recipe(dat_recipe) |> 
  add_model(lm_model)

# Fit the model
lm_fit <- lm_workflow |>
  fit(data = trees)

# Get the coefficients
trees_tidymodels_coefficients <- lm_fit |>
  tidy()

knitr::kable(trees_tidymodels_coefficients, digits = 4)
```

The fitted model is
$$
\hat{y} = -57.988 + 4.708x_1 + 0.339x_2
$$

where $x_1$ is `Girth` and $x_2$ is `Height`.

For every one-inch increase in Girth, the average Volume increases by 4.708 cubic feet, keeping Height fixed.

For every one-foot increase in Height, the average Volume increases by 0.339 cubic feet, keeping Girth fixed.


```{r warning=FALSE, message=FALSE}
# Get the global F-test p-value
trees_global_test <- lm_fit |>
  glance()

trees_global_test |>
  select(r.squared, statistic, p.value, df, df.residual) |>
  knitr::kable(digits = 4)

# Get the ANOVA table
# Note that SSR is split into individual predictors
trees_sequential_anova <- lm_fit |>
  extract_fit_engine() |>
  anova()

knitr::kable(trees_sequential_anova, digits = 4)
```


Since the global F-test p-value is very small, there is sufficient evidence to conclude that at least one slope coefficient is not 0.

The `anova()` table shown by R splits the regression sum of squares across predictors in the order they enter the model. The global F-test reported by `glance()` tests the predictors collectively.

:::

::: {.callout-note collapse="false" title="Global F-test versus sequential ANOVA table"}

The global $F$-test and the `anova()` table from `anova(lm_object)` answer related but different questions.

| Output | Main question | Important detail |
|---|---|---|
| `glance()` global F-test | Does the full model improve on the intercept-only model? | Tests all slopes together. |
| `anova(lm_object)` sequential ANOVA table | How much sum of squares is added by each predictor as it enters the model? | The order of predictors matters. |

For example, compare these two model formulas:

```{r warning=FALSE, message=FALSE}
anova_girth_first <- anova(lm(Volume ~ Girth + Height, data = trees))
anova_height_first <- anova(lm(Volume ~ Height + Girth, data = trees))

sequential_anova_compare <- bind_rows(
  as_tibble(anova_girth_first, rownames = "term") |>
    mutate(order = "Girth entered first"),
  as_tibble(anova_height_first, rownames = "term") |>
    mutate(order = "Height entered first")
) |>
  select(order, term, Df, `Sum Sq`, `Mean Sq`, `F value`, `Pr(>F)`)

knitr::kable(sequential_anova_compare, digits = 4)
```

The total model fit is the same either way, but the sequential ANOVA table allocates explained variation according to the order in which predictors enter the formula. This is why the global F-test is the cleaner answer to the question, "Is the predictor set useful as a group?"

:::

### A Bridge to Adjusted $R^2$

The ANOVA decomposition also connects directly to $R^2$.

Recall that

$$
R^2=\frac{SSR}{SSTO}=1-\frac{SSE}{SSTO}.
$$

This measures the proportion of sample variation in the response explained by the fitted model. However, ordinary $R^2$ never decreases when predictors are added, even if the new predictors are not useful. Adjusted $R^2$ adds a degrees-of-freedom penalty:

$$
R^2_{adj}=1-\frac{SSE/(n-p)}{SSTO/(n-1)}.
$$

Notice that this formula uses the same ingredients as the ANOVA table: $SSE$, $SSTO$, $n-p$, and $n-1$.

::: {.example title="Worked example: Connecting ANOVA to adjusted `R^2`"}

For the `trees` model, we can calculate $R^2$ and adjusted $R^2$ directly from the sums of squares.

```{r warning=FALSE, message=FALSE}
r_squared_trees <- ssr_trees / ssto_trees
adjusted_r_squared_trees <- 1 - (sse_trees / df_error) / (ssto_trees / df_total)

rsq_compare <- tibble(
  quantity = c("R-squared from sums of squares", "Adjusted R-squared from ANOVA pieces", "R-squared from lm summary", "Adjusted R-squared from lm summary"),
  value = c(
    r_squared_trees,
    adjusted_r_squared_trees,
    summary(trees_anova_fit)$r.squared,
    summary(trees_anova_fit)$adj.r.squared
  )
)

knitr::kable(rsq_compare, digits = 4)
```

Adjusted $R^2$ is useful because it rewards explained variation but penalizes the model for using more predictors. It is not a hypothesis test, but it is closely connected to the same variation-decomposition ideas behind ANOVA.

:::

## Recap

This chapter introduced the ANOVA approach to testing the overall utility of a multiple regression model.

| Idea | Meaning |
|---|---|
| **Global F-test** | Tests whether the predictors are collectively useful for explaining variation in the response. |
| **Null hypothesis** | $H_0:\beta_1=\beta_2=\cdots=\beta_{p-1}=0$. |
| **Alternative hypothesis** | At least one slope coefficient is not 0. |
| **SSTO** | Total variation in the response: variation around $\bar y$. |
| **SSR** | Regression variation: variation explained by the fitted model. |
| **SSE** | Error variation: variation left in the residuals. |
| **Sum-of-squares decomposition** | $SSTO=SSR+SSE$. |
| **Degrees-of-freedom decomposition** | $n-1=(p-1)+(n-p)$. |
| **MSR** | $MSR=SSR/(p-1)$, explained variation per regression degree of freedom. |
| **MSE** | $MSE=SSE/(n-p)$, unexplained variation per error degree of freedom. |
| **F statistic** | $F^*=MSR/MSE$. |
| **Large F statistic** | Evidence against $H_0$ and in favor of at least one useful predictor. |
| **Sequential ANOVA** | Splits SSR across predictors in formula order; useful, but order-dependent. |
| **$R^2$** | $R^2=SSR/SSTO=1-SSE/SSTO$, the proportion of sample response variation explained by the model. |
| **Adjusted $R^2$** | $R^2_{adj}=1-\dfrac{SSE/(n-p)}{SSTO/(n-1)}$, an $R^2$-like measure with a degrees-of-freedom penalty. |
| **Important limitation** | A significant global F-test does not identify which predictor is useful or prove that assumptions hold. |

## Check your understanding

::: {.callout-note collapse="false" title="Problems"}

1. What is the purpose of the global ANOVA $F$-test in multiple regression?

2. What null hypothesis is tested by the global $F$-test?

3. Why is conducting many individual $t$-tests not the same as conducting one global model test?

4. What does SSTO measure?

5. What does SSR measure?

6. What does SSE measure?

7. How are SSTO, SSR, and SSE related?

8. What are the degrees of freedom for regression, error, and total in a model with $p$ parameters and $n$ observations?

9. Why do we divide sums of squares by degrees of freedom to form mean squares?

10. What does the statistic $F^*=MSR/MSE$ compare?

11. What does a large value of $F^*$ suggest?

12. If the global $F$-test is significant, does that tell us which predictor is useful? Explain.

13. Does a significant global $F$-test prove that the regression assumptions are satisfied?

14. Why is the global $F$-test best viewed as a test of collective usefulness?

15. Why can the sequential ANOVA table from `anova(lm_object)` change when the order of predictors changes?

16. How is $R^2$ related to the ANOVA sums of squares?

17. Why does adjusted $R^2$ include a degrees-of-freedom penalty?

:::

::: {.callout-tip collapse="true" title="Solutions"}

1. **To test overall model utility.** The global $F$-test asks whether the predictors, as a group, explain a meaningful amount of variation in the response.

2. **All slopes are 0.** The null hypothesis is $H_0:\beta_1=\beta_2=\cdots=\beta_{p-1}=0$.

3. **Multiple tests inflate error risk.** Many individual $t$-tests create many chances for a Type I error. The global $F$-test evaluates the predictors together in one test.

4. **Total response variation.** SSTO measures the total variation of the observed responses around their mean.

5. **Explained variation.** SSR measures the part of the response variation explained by the fitted regression model.

6. **Unexplained variation.** SSE measures the residual variation left after fitting the model.

7. **They decompose total variation.** In least squares regression with an intercept, $SSTO=SSR+SSE$.

8. **Regression, error, total.** The regression degrees of freedom are $p-1$, the error degrees of freedom are $n-p$, and the total degrees of freedom are $n-1$.

9. **To put variation on a per-degree-of-freedom scale.** Mean squares compare sums of squares after accounting for their degrees of freedom.

10. **Explained variation relative to residual variation.** The $F$ statistic compares MSR to MSE.

11. **Evidence against the null.** A large $F^*$ suggests that the model explains more variation than expected if all slope coefficients were 0.

12. **No.** A significant global $F$-test says at least one slope is not 0, but it does not identify which one.

13. **No.** The test assumes the regression assumptions are reasonable; it does not verify them.

14. **It evaluates the predictor set.** The test compares a model with all predictors to an intercept-only model, so it addresses whether the predictors are collectively useful.

15. **Sequential means order-dependent.** The table gives each predictor credit for the extra sum of squares it explains after earlier predictors have already entered the model. If the order changes, that allocation can change.

16. **It is explained variation divided by total variation.** $R^2=SSR/SSTO=1-SSE/SSTO$.

17. **To discourage adding useless predictors.** Ordinary $R^2$ cannot decrease when predictors are added. Adjusted $R^2$ compares MSE to the total mean square, so it accounts for both unexplained variation and the number of parameters estimated.

:::









