---
editor: 
  markdown: 
    wrap: 72
---

# Model Assumptions and the ANOVA F-test

> "Regression analysis is the hydrogen bomb of the statistics
> arsenal." - Charles Wheelan

The steps we followed in developing a simple linear model are applicable
to the multiple regression model.

**Analyzing a Multiple Regression Model**

1.  Collect the *sample data* (i.e., the value of $y$, $x_1$, $x_2$,
    ..., $x_k$) for each unit in the sample.
2.  Hypothesize the *form* of the model (i.e., the deterministic
    component), $E(y)$. This involves choosing which independent
    variables to include in the model.
3.  Use the method of least squares to *estimate* the unknown
    *parameters*.
4.  Specify the probability distribution of the random error component
    $\varepsilon$ and estimate its *variance* $\sigma^2$.
5.  Statistically evaluate the *utility* of the model.
6.  Check that the *assumptions* on $\varepsilon$ are satisfied and make
    model modifications, if necessary.
7.  Finally, if the model is deemed adequate, use the fitted model to
    estimate the *mean* value of $y$ or to *predict* a particular value
    of $y$ for given values of the independent variables, and to make
    other inferences.

We will first assume the model is known and the assumptions hold. We
will then come back to step 2 and step 6 later. Let's first discuss what
the assumptions are.

## Model Assumptions

The assumptions for the multiple regression model are the same as the
simple linear model. That is,

-   *Linearity* - We assume the model is linear in the *parameters* but
    not necessarily linear in the *predictor* variables. Thus, some of
    the predictor variables may need to be transformed.
-   *constant* *variance*
-   *normality*
-   *independence*

The inferences on the multiple regression model will depend on these
assumptions holding. We will discuss how to check these later.

## A First-Order Model with Quantitative Predictors

Recall that a *first-order* model means that none of the predictor
variables are functions of any other predictor variables.

When the independent variables are quantitative, the $\beta$ parameters
in the first-order model have similar interpretations as the simple
regression model. The difference is that when we interpret the $\beta$
that multiplies one of the variables (e.g., $x_1$), we must be certain
to hold the values of the remaining independent variables (e.g., $x_2$,
$x_3$) *fixed*.

## Testing the Utility of a Model: The Analysis of Variance $F$-Test

The objective of step 5 in a multiple regression analysis is to conduct
a test of the *utility* of the model—that is, a test to determine
whether the model is adequate for predicting $y$.

Later, we will examine how to conduct $t$-tests on each $\beta$
parameter in a model, where $$
\begin{align*}
    H_0: \beta_j = 0, \qquad j=1, 2, \ldots, p-1
\end{align*}
$$

However, this approach is generally not a good way to determine whether
the *overall* model is contributing information for the prediction of
$y$. If we were to conduct a series of $t$-tests to determine whether
the independent variables are contributing to the predictive
relationship, we would be very likely to make one or more errors in
deciding which terms to retain in the model and which to exclude.

Suppose you fit a first-order model with 10 quantitative independent
variables, $x_1, x_2,..., x_{10}$, and decide to conduct $t$-tests on
all 10 individual $\beta$’s in the model, each at $\alpha = .05$.

For any one test, the probability of making a type I error is 0.05: $$
\begin{align*}
    P(\text{Reject } H_0|\beta_j=0) &= {0.05}\\
    & {= 1 - 0.95}
\end{align*}
$$ If we were to do ten of these tests (one for each predictor
variable), then the probability that at least one is a type I error is
$$
\begin{align*}
    P(\text{Reject at least one } H_0|\beta_1=\beta_2=\cdots=\beta_{p-1}=0)
    & = 1-[(1-\alpha)^{10}]\\
    & {= 1-(0.95)^{10}}\\
    &{ = 0.401}
\end{align*}
$$

Even if all the $\beta$ parameters (except $\beta_0$) in the model are
equal to 0, approximately 40% of the time you will incorrectly reject
the null hypothesis at least once and conclude that some $\beta$
parameter is nonzero. In other words, the overall Type I error is about
.40, not .05.

To illustrate this inflated type I error, let's look at the following
example.

::: {#exm-12_1}
We will simulate a sample of size 200 with a response variable $y$ and
ten predictor variables $x_1, x_2, \ldots, x_{10}$. The random error
term $\varepsilon$ will be a standard normal random variable.

The true model will have each coefficient $\beta$ set to 0, with the
exception of $\beta_0$ which will be 20. We will conduct a $t$-test for
each coefficient (except $\beta_0$). Since the true coefficient is 0, we
expect to see p-values greater than 0.05 for each coefficient.

We will fit the model and test the coefficients 1000 times. We will
count the number of times at least one of the coefficients was less than
0.05 (which would lead to a Type I error).

```{r warning=FALSE, message=FALSE}
library(tidyverse)

set.seed(3430)

n = 200

type1error = numeric(1000)

for(i in 1:1000){
  
  x = runif(n*10) |> matrix(ncol = 10)
  
  eps = rnorm(n)
  
  y = 20 + 0*x[,1] + 0*x[,2] + 0*x[,3] + 0*x[,4] +
    0*x[,5] + 0*x[,6] + 0*x[,7] + 0*x[,8] + 0*x[,9] +
    0*x[,10] + eps
  
  dat = tibble(y, x1 = x[,1], x2 = x[,2], x3 = x[,3],
               x4 = x[,4], x5 = x[,5], x6 = x[,6], x7 = x[,7],
               x8 = x[,8], x9 = x[,9], x10 = x[,10],)
  
  fit = lm(y~., data = dat)
  
  type1error[i] = any(summary(fit)$coefficients[-1,4] < 0.05)
}

mean(type1error)

```

We see that 41.4% of the time, we had at least one coefficient have a
p-value smaller than 0.05. This corresponds to the probability of at
least one Type I error of around 0.401.
:::

In multiple regression models for which a large number of independent
variables are being considered, conducting a series of $t$-tests may
cause the experimenter to include a large number of insignificant
variables and exclude some useful ones. If we want to test the utility
of a multiple regression model, we will need a *global* *test* (one that
encompasses all the $\beta$ parameters).

### Partitioning the Sum of Squares

Recall when we discussed the coefficient of determination that we used
$SS_{yy}$ to denote the variability of the response variable $y$ from
its mean $\bar y$ (without regard to the model involving $x$).

Another name for $SS_{yy}$ is the *sum of squares total* (SSTO). We call
it this since it gives us a measure of total variability in $y$.

In multiple regression, SSTO is still the same as $SS_{yy}$. In matrix
notation this can be expressed as $$
\begin{align}
SSTO & ={\bf Y}^{\prime}{\bf Y}-\left(\frac{1}{n}\right){\bf Y}^{\prime}{\bf J}{\bf Y}
\end{align}
$$ {#eq-w4-34}

#### Sum of Square Error (SSE)

Also recall the variability of $y$ about the regression line (in simple
linear regression) was expressed by SSE. We can think of this as the
variability of $y$ remaining after exampling some of the variability
with the regression model.

In multiple regression, SSE is still the sum of the square distances
between the response $y$ and the fitted model $\hat{y}$. Now, the fitted
model is the fitted *hyperplane* instead of a line.

The SSE can be expressed in matrix terms as $$
\begin{align}
SSE & =\left({\bf Y}-{\bf X}{\bf b}\right)^{\prime}\left({\bf Y}-{\bf X}{\bf b}\right)\\
& ={\bf Y}^{\prime}{\bf Y}-{\bf b}^{\prime}{\bf X}^{\prime}{\bf Y}
\end{align}
$$ {#eq-w3_35}

#### Sum of Square Regression (SSR)

If SSTO is the total variability of $y$ (without of regard to the
predictor variables), and SSE is the variability of $y$ left over after
explaining the variability of $y$ with the model (including the
predictor variables), we might want to know the variability of $y$
explained by the model.

We call the variability explained by the regression model the *sum of
squares regression* (SSR).

We will show below that SSR can be expressed as $$
\begin{align}
SSR & =\sum\left(\hat{y}_{i}-\bar{y}\right)^{2}
\end{align}
$$ {#eq-w4_36} which can be expressed in matrix terms as $$
\begin{align}
SSR & ={\bf b}^{\prime}{\bf X}^{\prime}{\bf Y}-\left(\frac{1}{n}\right){\bf Y}^{\prime}{\bf J}{\bf Y}
\end{align}
$$ {#eq-w4_37}

#### Components of SSTO

To see how SSTO, SSR, and SSE relate to each other, consider how SSTO is
a sum of squares of $y$ from its mean $\bar{y}$: $$
\begin{align*}
y_{i}-\bar{y}
\end{align*}
$$

We can add and subtract the fitted value $\hat{y}_{i}$ to get $$
\begin{align*}
y_{i}-\bar{y} & =y_{i}-\hat{y}_{i}+\hat{y}_{i}-\bar{y}\\
& =\left(y_{i}-\hat{y}_{i}\right)+\left(\hat{y}_{i}-\bar{y}\right)
\end{align*}
$$

Squaring both sides gives us $$
\begin{align*}
\left(y_{i}-\bar{y}\right)^{2} & =\left[\left(y_{i}-\hat{y}_{i}\right)+\left(\hat{y}_{i}-\bar{y}\right)\right]^{2}\\
& =\left(y_{i}-\hat{y}_{i}\right)^{2}+\left(\hat{y}_{i}-\bar{y}\right)^{2}+2\left(y_{i}-\hat{y}_{i}\right)\left(\hat{y}_{i}-\bar{y}\right)
\end{align*}
$$

Summing both sides gives us $$
\begin{align*}
\sum\left(y_{i}-\bar{y}\right)^{2} & =\sum\left(y_{i}-\hat{y}_{i}\right)^{2}+\sum\left(\hat{y}_{i}-\bar{y}\right)^{2}+2\sum\left(y_{i}-\hat{y}_{i}\right)\left(\hat{y}_{i}-\bar{y}\right)\\
& =\sum\left(y_{i}-\hat{y}_{i}\right)^{2}+\sum\left(\hat{y}_{i}-\bar{y}\right)^{2}+2\sum\hat{y}_{i}e_{i}-2\bar{y}\sum e_{i}
\end{align*}
$$

Note that $\sum\hat{y}_{i}e_{i}=0$ 
<!-- from @eq-w1_22  -->
and $\sum e_{i}=0$.
<!-- from @eq-w1_20. -->

Therefore, we have $$
\begin{align*}
\sum\left(y_{i}-\bar{y}\right)^{2} & =\sum\left(\hat{y}_{i}-\bar{y}\right)^{2}+\sum\left(y_{i}-\hat{y}_{i}\right)^{2}\\
SSTO & =SSR+SSE 
\end{align*}
$$ {#eq-4_38}

We call this the *decomposition* of SSTO.

#### Degrees of Freedom

The degrees of freedom can be decomposed as well. Note that the degrees
of freedom for SSTO is $$
\begin{align*}
df_{SSTO} & =n-1
\end{align*}
$$ since the mean of $y$ is needed to be estimated with $\bar{y}$.

The degrees of freedom for SSE is $$
\begin{align*}
df_{SSE} & =n-p
\end{align*}
$$ since the $p$ coefficients $\beta_{0},\ldots,\beta_{p-1}$ need to be
estimated with $\hat{\beta}_{0},\ldots,\hat{\beta}_{p-1}.$

For SSR, the degrees of freedom is $$
\begin{align*}
df_{SSR} & =p-1
\end{align*}
$$ since there are $p$ estimated coefficients
$\hat{\beta}_{0},\ldots,\hat{\beta}_{p-1}$ but need to estimate the mean
of $y$ with $\bar{y}$.

Decomposing the degrees of freedom give us $$
\begin{align}
n-1 & =p-1+n-p\\
df_{SSTO} & =df_{SSR}+df_{SSE}
\end{align}
$$ {#eq-w4_29}

### The Analysis of Variance (ANOVA) Table

The sums of squares and degrees of freedom are commonly displayed in an
analysis of variance (ANOVA) table:

|   Source   |     df      |   SS   | MS  |  F  | p-value |
|:----------:|:-----------:|:------:|:---:|:---:|:-------:|
| Regression | $df_{SSR}$  | $SSR$  |     |     |         |
|   Error    | $df_{SSE}$  | $SSE$  |     |     |         |
|   Total    | $df_{SSTO}$ | $SSTO$ |     |     |         |



#### Mean Squares

Recall that if we divide SSE by its degrees of freedom, we obtain the
mean square error: 
$$
\begin{align}
MSE & =\frac{SSE}{n-p}
\end{align}
$${#eq-w4_40}


Likewise, if we divide SSR by its degrees of freedom, we obtain the
*mean square regression*: 
$$
\begin{align}
MSR & =\frac{SSR}{p-1}
\end{align} 
$${#eq-w4_41}

These values are also included in the ANOVA table:

|   Source   |     df      |   SS   | MS  |  F  | p-value |
|:----------:|:-----------:|:------:|:---:|:---:|:-------:|
| Regression | $df_{SSR}$  | $SSR$  |  $MSR$   |     |         |
|   Error    | $df_{SSE}$  | $SSE$  | $MSE$    |     |         |
|   Total    | $df_{SSTO}$ | $SSTO$ |     |     |         |


Note that although the sum of squares and degrees of freedom decompose,
the mean squares do not. That is 
$$
\begin{align*}
\frac{SSTO}{n-1} & \ne MSR+MSE
\end{align*}
$$

In fact, the mean square for the total ($SSTO/n-1$) does
not usually show up on the ANOVA table.

### The ANOVA F-test

We tested the slope in simple regression to see if there is a
significant linear relationship between $X$ and $Y$.

In multiple regression, we will want to see if there is any significant
linear relationship between any of the $x$s and $y$. Thus, we want to
test the hypotheses 
$$
\begin{align*}
H_{0}: & \beta_{1}=\beta_{2}=\cdots=\beta_{p-1}=0\\
H_{a}: & \text{at least one } \beta \text{ is not equal to zero}
\end{align*}
$$

To construct a test statistic, we first note that 
$$
\begin{align*}
\frac{SSE}{\sigma^{2}} & \sim\chi^{2}\left(n-p\right)
\end{align*}
$$
Also, if $H_{0}$ is true, then 
$$
\begin{align*}
\frac{SSR}{\sigma^{2}} & \sim\chi^{2}\left(p-1\right)
\end{align*}
$$

The ratio of two independent chi-square random variables divided by
their degrees of freedom give a statistic that follows a F-distribution.

Since $SSE/\sigma^{2}$ and $SSR/\sigma^{2}$ are independent (proof not
give here), then under $H_{0}$, we can construct a test statistic as
$$
\begin{align}
F^{*} & =\left(\frac{\frac{SSR}{\sigma^{2}}}{p-1}\right)\div\left(\frac{\frac{SSE}{\sigma^{2}}}{n-p}\right)\\
& =\left(\frac{SSR}{p-1}\right)\div\left(\frac{SSE}{n-p}\right)\\
& =\frac{MSR}{MSE}
\end{align}
$${#eq-w4_42}


Large values of $F^{*}$ indicate evidence for $H_{a}$.




The test statistic and p-value are the last two components of the ANOVA
table:

|   Source   |     df      |   SS   | MS  |  F  | p-value |
|:----------:|:-----------:|:------:|:---:|:---:|:-------:|
| Regression | $df_{SSR}$  | $SSR$  |  $MSR$   | $F^{*}$    |  $P\left(Z\ge Z^{*}\right)$       |
|   Error    | $df_{SSE}$  | $SSE$  | $MSE$    |     |         |
|   Total    | $df_{SSTO}$ | $SSTO$ |     |     |         |





::: {#exm-12_2}
We will fit a multiple regression model to the `trees` dataset. The tidymodels framework will be used. 

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(tidymodels)

data(trees)

#prepare data
dat_recipe = recipe(Volume ~ Girth + Height, data = trees) 

#setup model
lm_model = linear_reg() |>
  set_engine("lm")

#setup the workflow
lm_workflow = workflow() |>
  add_recipe(dat_recipe) |> 
  add_model(lm_model)

#fit the model
lm_fit = lm_workflow |>
  fit(data = trees)

#to get the coefficients
lm_fit |> tidy()
```

The fitted model is
$$
\hat{y} = -57.988 + 4.708x_1 + 0.339x_2
$$
For every one inch increase in Girth, the average Volume increases by 4.708 cubit ft, keeping Height fixed. 

For every one ft increase in Height, the average Volume increases by 0.339 cubit ft, keeping Girth fixed. 


```{r warning=FALSE, message=FALSE}
#to get the global F-test p-value
lm_fit |> glance()

#To get the ANOVA table
#Note that SSR is split into individual predictors
lm_fit |> extract_fit_engine() |> anova()
```


Since the p-value of $1.071\times 10^{-18}$ is less than 0.05, there is sufficient evidence to conclude that at least one of the coefficients is not zero.

:::









