

# Ridge Regression and the LASSO {#sec-ridge}

> "All generalizations are false, including this one." - Mark Twain



## Ridge Regression

We have seen that multicollinearity causes the least
squares estimates to be *imprecise*. 




If we allow some bias in our estimators, then we can have estimators
that are more precise. This is due to the relationship 
$$
\begin{align*}
E\left[\left(\hat{\beta}-\beta\right)^{2}\right] & =Var\left[\hat{\beta}\right]+\left(E\left[\hat{\beta}\right]-\beta\right)^{2}
\end{align*}
$$

*Ridge* regression starts by transforming the variables as
$$
\begin{align}
Y_{i}^{*} & =\frac{1}{\sqrt{n-1}}\left(\frac{Y_{i}-\overline{Y}}{s_{Y}}\right)\nonumber\\
X_{ik}^{*} & =\frac{1}{\sqrt{n-1}}\left(\frac{X_{ik}-\overline{X}_{k}}{s_{k}}\right)\label{eq:w5_23}
\end{align}
$$
This is known as the *correlation* *transformation*. 




The estimates are then found by *penalizing* the least squares criterion
$$
\begin{align}
Q & =\sum\left[Y_{i}^{*}-\left(b_{1}^{*}X_{i1}^{*}+\cdots+b_{p-1}^{*}X_{i,p-1}^{*}\right)\right]^{2}+\lambda\left[\sum_{j=1}^{p-1}\left(b_{j}^{*}\right)^{2}\right]
\end{align}
$${#eq-w5_24}

The added term gives a penalty for large coefficients. Thus, the estimators
are biased toward 0. Because of this, the ridge estimators are called
*shrinkage* estimators. 




There are a number of methods for determining $\lambda$ which are
beyond the scope of our class. In R, we will use the `glmnet` engine to fit the model. 




The ridge estimators are more stable than the ols estimators, however, the distributional properties of these estimators are not easily found. 


::: {#exm-19_1}

## Ridge Regression for `mtcars`

Let's examine the `mtcars` data last seen in @exm-16_3. Recall that `disp`, `hp`, and `wt` need to be log-transformed to make the relationships linear.

We will now use the "glmnet" engine for ridge regression. We set the value of $\lambda$ in @eq-w5_24 with the `penalty` argument in `linear_reg`. We will just arbitrarily use 0.1 for now. We will later in this section discuss how to choose a value of `penalty`. 


```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(tidymodels)

dat_recipe = recipe(mpg ~ disp + hp + drat + wt + qsec, data = mtcars) |> 
  step_mutate(
    log_disp = log(disp),
    log_hp = log(hp),
    log_wt = log(wt)
  ) |> 
  step_rm(disp, hp, wt) |> 
  step_normalize(all_numeric_predictors())

#penalty is the lambda hyperparameter
#mixture = 0 indicates ridge regression
lm_model = linear_reg(mixture = 0, penalty = 0.1) |>
  set_engine("glmnet")

wf = workflow() |> 
  add_recipe(dat_recipe) |> 
  add_model(lm_model)

fit = wf |> fit(mtcars)

fit |> tidy()

```

We see that `log_wt` has the biggest impact on `mpg` since it has the largest coefficient in magnitude. The negative sign just indicates that the linear relationship between `log_wt` and `mpg` is negative. 

We also see that `drat` has the smallest impact on `mpg` given the other variables are in the model, followed by `qsec` for the second smallest impact. Let's reexamine the scatterplot matrix to see if this make sense.

```{r warning=FALSE, message=FALSE}
library(GGally)

mtcars |>
  mutate(log_disp = log(disp),
         log_hp = log(hp),
         log_wt = log(wt)) |> 
  select(mpg, drat, qsec, log_disp, log_hp, log_wt) |> 
  ggpairs()


```

Looking at the plots of `mpg` versus the other variables, we see that a linear relationship is apparent for all of the predictors. Between `drat` and `qsec`, `drat` appears stronger ($r=0.681$) than `qsec` ($r=0.419$). But why did ridge regression show us that `qsec` has a larger impact on `mpg` than `drat`? Ridge regression shows the impact given the other variables are included in the model. Looking at the scatterplot matrix, `drat` is more correlated with the other predictor variables than `qsec`. Thus, when every other predictor is included, `drat` does not explain much more variability in `mpg`. Thus, it is not as important. 

:::

### The effect of the `penalty`

In @exm-19_1, we chose the `penalty` ($\lambda$) arbitrarily to be 0.1. Are there other values of `penalty` that would fit the data better?  

We can try different penalties using the `tidy` function:

```{r warning=FALSE, message=FALSE}
fit |> tidy(penalty = .75)

```

We can even plot the value of each coefficient for different penalties.

```{r warning=FALSE, message=FALSE}

fit |> extract_fit_parsnip() |> autoplot()

```

We see that for some penalties, the importance of some coefficients change. 

Let's use cross-validation to help determine the `penalty`.

### Cross-Validation


Cross-validation is a powerful resampling method used to evaluate the performance of models while minimizing bias and variance. Within the tidymodels framework, this technique ensures that models generalize well to unseen data by fitting and validating on different subsets of the dataset.

#### Types of Cross-Validation

1. K-Fold Cross-Validation:

   - The data is divided into *k* equal-sized folds (or partitions).
   - The model is fit on $k - 1$ folds and validated on the remaining fold.
   - This process is repeated *k* times, with each fold serving as the validation set once.
   - Final performance is computed by averaging metrics across all folds.

2. Repeated K-Fold Cross-Validation:

   - This variant involves running k-fold cross-validation multiple times with different splits, which provides more reliable estimates by averaging over several repetitions.

3. Leave-One-Out Cross-Validation (LOOCV):

   - Each observation in the dataset serves as the validation set exactly once. This method can be computationally expensive but works well for small datasets.
   

#### Implementing Cross-Validation in tidymodels

To implement cross-validation, the `rsample` package (part of tidymodels) provides essential tools for splitting the data. Below is an example of performing **5-fold cross-validation**:

```{r warning=FALSE, message=FALSE}
set.seed(1004) #to reproduce results


cv_folds = vfold_cv(mtcars, v = 5)
```



#### Using Cross-Validation with Workflows

Within tidymodels, you can streamline model fitting with workflows and use `fit_resamples()` to fit models on the resamples created by cross-validation:

```{r warning=FALSE, message=FALSE}
dat_recipe = recipe(mpg ~ disp + hp + drat + wt + qsec, data = mtcars) |> 
  step_mutate(
    log_disp = log(disp),
    log_hp = log(hp),
    log_wt = log(wt)
  ) |> 
  step_rm(disp, hp, wt) |> 
  step_normalize(all_numeric_predictors())

lm_model = linear_reg(mixture = 0, penalty = 0.1) |>
  set_engine("glmnet")

wf = workflow() |> 
  add_recipe(dat_recipe) |> 
  add_model(lm_model)

fits = wf |> fit_resamples(wf, resamples = cv_folds)

# Collect and summarize performance metrics
collect_metrics(fits)
```

### Tune the Penalty

We can now have tidymodels try different values of `penalty` and determine what is the best by checking the performance on the folded data. Using the folded data helps us determine what value of `penalty` provides the best fit but also what value does the best job at predicting data that was not used to determine the `penalty`. 

We need to setup some values of `penalty` to try. These values can be constructed using `grid_regular`. After constructing the grid of possible values of `penalty` to try, we can then have tidymodels do all of the fitting using `tune_grid`.

We will put all of these pieces together in the following example.



::: {#exm-19_2}

## Fine tune the penalty for `mtcars`

```{r warning=FALSE, message=FALSE}

set.seed(1004) #to reproduce results

#prepare the recipe
dat_recipe = recipe(mpg ~ disp + hp + drat + wt + qsec, data = mtcars) |> 
  step_mutate(
    log_disp = log(disp),
    log_hp = log(hp),
    log_wt = log(wt)
  ) |> 
  step_rm(disp, hp, wt) |> 
  step_normalize(all_numeric_predictors())

#setup the folded data, let's try 5 folds
cv_folds = vfold_cv(mtcars, v = 5)

#setup the model and set penalty to tune
lm_model = linear_reg(mixture = 0, penalty = tune()) |>
  set_engine("glmnet")

wf = workflow() |> 
  add_recipe(dat_recipe) |> 
  add_model(lm_model)

#setup possible values of penalty
#this will try 100 different values of penalty
penalty_grid = grid_regular(penalty(range=c(-6, 4)), levels = 100)

#now tune the grid with everthing we just set up
tune_model = tune_grid(
  wf,
  resamples = cv_folds, 
  grid = penalty_grid,
  metrics = metric_set(rmse, rsq)
)


```

At the point, differing values of `penalty` has been tried as determined by `grid_regular`. We can view the results in a few ways. We can plot the mean metric for each value of the penalty, or we can just ask R to show the top few penalties based on either "rmse" or "rsq". 

```{r warning=FALSE, message=FALSE}
#plot the metrics for different values of the penalty
tune_model |>   autoplot()

#find the best penalty in terms of rmse
tune_model |> show_best(metric = "rmse")

#find the best penalty in terms of rsq
tune_model |> show_best(metric = "rsq")

```

Note the "best" penalty value differs from "rmse" to "rsq". Looking at the plot above, we can see there is not much difference in either metric for the "best" penalty value for either case. 

Let's now get the best penalty from one of the metrics. We will use "rsq" here.


```{r warning=FALSE, message=FALSE}

best_penalty = tune_model |> select_best(metric = "rsq")
best_penalty

```

With this best penalty, we can finalize the workflow to use this penalty for the ridge regression.

```{r warning=FALSE, message=FALSE}
final_wf = finalize_workflow(wf, best_penalty)

final_fit = fit(final_wf, data = mtcars)

final_fit |> tidy()

```
:::



## Lasso


### Shrinkage as a Variable Selector




In ridge regression, the estimates *shrink* to zero for predictors that do not have a significant linear relationship on $Y$ given the other variables. 





The coefficient estimates shrink to zero, but do not equal zero.



The lasso (*least* *absolute* *shrinkage* and *selection* *operator*) is a shrinkage method like ridge, with subtle but important differences.

                            



The lasso estimate is defined by
$$
\begin{align}
Q_{lasso} & =\sum_{i=1}^{n}\left(Y_{i}-\left(\beta_{0}+\beta_{1}X_{1i}+\cdots+\beta_{p-1}X_{p-1,i}\right)\right)^{2}+\lambda\sum_{j=1}^{p-1}\left|\beta_{j}\right| 
\end{align}
$${#eq-w7_9}


Similarly to ridge regression, the value of $\lambda$ determines how biased the estimates will be. 

As $\lambda$ increases, the beta coefficients shrink toward zero with the *least* associated beta coefficients decreasing all the way to 0 before the more strongly associated beta coefficients. 


As a result, numerous beta coefficients that are not strongly associated with the outcome are decreased to *zero*, which is equivalent to removing those variables from the model. 





In this way, the lasso can be used as a *variable* *selection* method. 



::: {#exm-19_3}

## Lasso for `mtcars`

The procedure to fit the lasso is very similar to ridge regression. The only change is in the `mixture` argument in `linear_reg`. We will set this to 1 to do lasso.

```{r warning=FALSE, message=FALSE}

set.seed(1004) #to reproduce results

#prepare the recipe
dat_recipe = recipe(mpg ~ disp + hp + drat + wt + qsec, data = mtcars) |> 
  step_mutate(
    log_disp = log(disp),
    log_hp = log(hp),
    log_wt = log(wt)
  ) |> 
  step_rm(disp, hp, wt) |> 
  step_normalize(all_numeric_predictors())

#setup the folded data, let's try 5 folds
cv_folds = vfold_cv(mtcars, v = 5)

#setup the model and set penalty to tune
lm_model = linear_reg(mixture = 1, penalty = tune()) |>
  set_engine("glmnet")

wf = workflow() |> 
  add_recipe(dat_recipe) |> 
  add_model(lm_model)

#setup possible values of penalty
#this will try 100 different values of penalty
penalty_grid = grid_regular(penalty(range=c(-4, 4)), levels = 100)

#now tune the grid with everything we just set up
tune_model = tune_grid(
  wf,
  resamples = cv_folds, 
  grid = penalty_grid,
  metrics = metric_set(rmse, rsq)
)

tune_model |>   autoplot()

best_penalty = tune_model |> select_best(metric = "rsq")

final_wf = finalize_workflow(wf, best_penalty)

final_fit = fit(final_wf, data = mtcars)

final_fit |> tidy()


```

We see in this dataset, `drat` and `qsec` shrunk to 0. This indicates that they are not important to modeling `mpg`. The largest magnitude is `log_wt` which indicates that it has the largest impact on `mpg`

In the next example, we will fit a lasso model to the bodyfat data.

:::

::: {#exm-19_4}

## Lasso for the bodyfat data. 

```{r warning=FALSE, message=FALSE}

dat = read_table("bodyfat.txt")

#prepare the recipe
dat_recipe = recipe(bfat ~ tri + thigh + midarm, data = dat) |> 
  step_normalize(all_numeric_predictors())

#setup the folded data, let's try 5 folds
cv_folds = vfold_cv(dat, v = 5)

#setup the model and set penalty to tune
lm_model = linear_reg(mixture = 1, penalty = tune()) |>
  set_engine("glmnet")

wf = workflow() |> 
  add_recipe(dat_recipe) |> 
  add_model(lm_model)

#setup possible values of penalty
#this will try 100 different values of penalty
penalty_grid = grid_regular(penalty(range=c(-4, 4)), levels = 100)

#now tune the grid with everything we just set up
tune_model = tune_grid(
  wf,
  resamples = cv_folds, 
  grid = penalty_grid,
  metrics = metric_set(rmse, rsq)
)

tune_model |> autoplot()

best_penalty = tune_model |> select_best(metric = "rsq")

final_wf = finalize_workflow(wf, best_penalty)

final_fit = fit(final_wf, data = dat)

final_fit |> tidy()


```

We see that `tri` has the biggest impact on `bfat`. The predictor `thigh` shrunk down to 0. Recall in previous examples when we used the bodyfat data that `tri` and `thigh` are highly correlated. 

When two predictor variables are highly correlated, the lasso will arbitrarily have one of those variables shrink to zero. The practitioner should understand that the one variable that was not selected could in fact be more appropriate for their model than the one that was selected. 

:::



