

# Residual Analysis and Remedial Measures

> "And I knew exactly what to do. But in a much more real sense, I had no idea what to do." - Michael Scott



We discussed checking the assumptions of a simple linear regression model in @sec-checking and @sec-checking2. For multiple regression, we have already discussed the linearity assumption in @sec-linearity and multicollinearity in @sec-multicollinearity. We will now revisit the remaining assumptions of the multiple regression model and present some remedial measures. 




## Assumptions for the Multiple Regression Model



For the normal errors multiple regression model in @eq-w4_18 we need to check the assumptions. 



We will use the residuals here as we did in simple regression. 



Let's begin by listing all of the assumptions for the model in @eq-w4_18: 


1. *Linearity*: There is a linear relationship between $y$ and each of the predictor variables. (see @sec-linearity)
2. *Normality* of the residuals: In the normal error model, the error terms should be approximately normally distributed. 
3.*Constant* variance: The variance of the error terms should be constant throughout the range of the predictor variables. 
4.*Independence* of the residuals: We assume the error terms are independent of each other.
5.*Uncorrelated* predictor variables: There is no multicollinearity present between the predictor variables (see @sec-multicollinearity). 




#3# Normality and Transformations

We can check normality as we did in simple regression by plotting the residuals in a *QQ plot* or by using the *Shapiro Wilk* test. 

                            



It is good practice to do both so that you can get a formal test and a visualization. 





If the data are not normally distributed, then a transformation on $y$, such as a *Box-Cox* transformation may be helpful. 






### Box-Cox Transformation


In the Box-Cox transformation, a procedure is used to determine a power ($\lambda$) of $y$
$$
\begin{align*}
Y^\prime =& \frac{Y^\lambda - 1}{\lambda} &\text{ if }\lambda\ne 0\\
Y^\prime =& \ln Y & \text{ if }\lambda= 0
\end{align*}
$$
that results in a regression fit with residuals as close to normality as possible. 


In tidymodels, we can conduct a Box-Cox transformation using `step_BoxCox` in the recipe.


::: {#exm-21_1}

## Handspan data

Let's look at the handspan data first examine in @exm-15_1. 

```{r message=FALSE, warning=F}
library(tidyverse)
library(tidymodels)
library(olsrr)

dat = read_csv("SurveyMeasurements.csv")

dat_recipe = recipe(`right handspan (cm)` ~ `height (inches)` + sex, 
                    data = dat) |> 
  step_naomit() |> 
  step_dummy(all_nominal_predictors()) 

lm_model = linear_reg() |>
  set_engine("lm")

lm_workflow = workflow() |>
  add_recipe(dat_recipe) |> 
  add_model(lm_model)

lm_fit = lm_workflow |>
  fit(data = dat)

lm_fit |> extract_fit_engine() |> ols_plot_resid_qq()



dat_recipe_bc = recipe(`right handspan (cm)` ~ `height (inches)` + sex, 
                    data = dat) |> 
  step_naomit() |> 
  step_dummy(all_nominal_predictors()) |> 
  step_BoxCox(`right handspan (cm)`)

lm_model = linear_reg() |>
  set_engine("lm")

lm_workflow = workflow() |>
  add_recipe(dat_recipe_bc) |> 
  add_model(lm_model)

lm_fit = lm_workflow |>
  fit(data = dat)

lm_fit |> extract_fit_engine() |> ols_plot_resid_qq()
```

We see that the transformation dampened the skewness of residuals but they are still nonnormal. The Box-Cox transformation does not guarantee the result will be normal, but is attempts to make it closer to normal. 

If we would like to see the value of $\lambda$ used in `step_BoxCox`, we can do so with the following:

```{r message=FALSE, warning=F}
#use number = 3 since step_BoxCox is the third step in the recipe
prep(dat_recipe_bc, training = dat) |> tidy(number=3)

```

:::



One downside to transformations on $y$ is that the *interpretability* on the confidence intervals may be difficult to grasp. 





For example, if we have a confidence interval for the mean response where $y$ was log-transformed  as $(.35, .78)$, we could back transform by doing the inverse function of log which is the exponential function. So we would have $(\exp(.35), \exp(.78)=(1.419, 2.181)$. However, this is not a confidence interval for the mean $y$, it is instead a confidence interval for the median $y$. 





Other back transformations may not have this interpretation. If the transformation is not monotonic, then the backtransformed confidence interval may not have the desired coverage. 





Thus, a transformation may not be the best option if the goal is to obtain confidence and prediction intervals for the response.




### Independence


We can visualize the correlation by using a *acf* plot. 


A formal test for significant autocorrelation is the *Breusch-Godfrey* Test as was used in simple linear regression. 


If significant autocorrelation is present, then a time series model is necessary (beyond the scope of this course).


### Constant Variance


We can visualize the variance of the residuals as we did in simple regression. A plot of the residuals versus the fitted values can be examined for the "cone" or "megaphone" shape. 

   

We can also use the *Breusch-Pagan* test as we did in simple regression to formally test for nonconstant variance.


## Remedial Measures


### Overview of Assumptions and Remedial Measures


We will list again the assumptions of model @eq-w4_18
and also discuss the problems violations of these assumptions present and what can be done when they are violated.


1. Linearity: Violating linearity will lead to incorrect conclusions about the relationship between $y$ and the predictor variables and thus lead to incorrect estimations and predictions.

    As discussed previously, a transformation on the $x$ variables can help with linearity. 
    
    
2. Normality of the residuals: The t-test, F-test, confidence intervals, and prediction intervals all depend on normality of the residuals (or at least symmetric distribution).

    If we do not have normality, then we cannot do the inferences for the model. A transformation on $y$ may help with normality (or at least get to a symmetric distribution), however, we tend to lose interpretation in our inferences.

    If we cannot obtain normality through a transformation, or we wish to keep interpretation and not do a transformation, then we can still do inferences based on *bootstrapping*. We will discuss bootstrapping in more detail below. 
    
3. Constant variance: Very much like the normality assumption, the constant variance assumption affects the inferences made for the model. Recall all the equations for the variance of the estimated coefficients (@eq-w4_31), the variance of the mean response (@eq-w4_47), and the variance of the prediction (@eq-w4_50). Each involves MSE which is an estimate of the constant variance $\sigma^2$. If $\sigma^2$ is not constant, then MSE is not the estimator we need. We would need an estimator that takes into account the nonconstant variance. 


    Taking a transformation of $y$ could help stabilize the variance, however, we would lose interpretation due to the transformation. Using a method known as weighted least squares (discussed below) can help with nonconstant variance.

4. Independence of the residuals: If there is significant autocorrelation in the residuals, then a time series model will be needed. 

    In simple regression, we discussed that a difference could be used to remove simple autocorrelation (see @sec-checking2). The downside to doing a difference is that, once again, you lose interpretation in your inferences. 

 

5. Uncorrelated predictor variables: Previously, we discussed the problems of having highly correlated predictor variables which is known as multicollinearity. 

    In some applications, it is very difficult to have a subset of predictor variables that do not have multicollinearity. When this is the case, we can use ridge regression or lasso. 



### Bootstrap Sampling

#### Repeated Sampling

If we were to repeatedly take a sample of size $n$ from the population of interest and then find the least squares estimates each time, we could plot these estimates to estimate their sampling distributions. This is known as *repeated sampling*.

```{r message=FALSE, warning=F, echo = F}
dat = read_csv("SurveyMeasurements.csv")
cb = choose(nrow(dat), 30)
# cb = format(cb, scientific = FALSE, big.mark = ',')


extract <- function(x){
	e <- ifelse(x == 0, 0, floor(log10(x)))
	m <- x/10^e
	list(mantissa = m, exponent = e)
}

cb = extract(cb)

```

For example, let's consider the handspan and height measurements from @exm-15_1. These measurements were from `r nrow(dat)` college students. Suppose these students are now the population of interest.


We will take a random sample of $n=30$ from this population of `r nrow(dat)` college students. We could look at all possible samples of size $n=30$ and the resulting least squares estimates $b_0$ and $b_1$. This would give us the exact sampling distribution for each.  In this example with a relatively small population size of `r nrow(dat)`, the number of possible samples of size 30 is
$$
\begin{align*}
\binom{`r nrow(dat)`}{30} & =`r cb$mantissa`\times 10^{`r cb$exponent`}
\end{align*}
$$


It would be infeasible to examine this many samples, however, we could look at enough samples (tens of thousands) to get an estimate of the sampling distributions. 


Let's first get a scatterplot all `r nrow(dat)` population measurements. They are colored gray. The least squares line for the entire population is shown in blue. 

```{r message=FALSE, warning=F}
dat = read_csv("SurveyMeasurements.csv")

p1 = dat |> ggplot(aes(x=`height (inches)`, y=`right handspan (cm)`)) +
  geom_point(color = "gray") +
  geom_smooth(method="lm", se=FALSE)

p1

```



Let's now take a random sample of size $n=30$. This sample will be shown as red points in the scatterplot below. The red line represents the fitted line for the sample. 

```{r message=FALSE, warning=F}
set.seed(34)#to replicate results

dat_sample = sample_n(dat, 30)

fit_sample = lm(`right handspan (cm)`~`height (inches)`, data = dat_sample)

p1 + geom_point(data=dat_sample, color = "red") +
  geom_smooth(data=dat_sample, method="lm", se=FALSE, color = "red")

```





The sample in the figure above is just one sample of size $n=30$. Suppose we repeated this sampling and got a total of 10,000 samples of size $n=30$ from the data. Each time, we fit the least squares line to the sample and save the y-intercept ($b_0$) and slope ($b_1$). 


The plots below show the histograms of the least squares estimates $b_0$ and $b_1$ from the repeated sampling. The blue and red vertical lines in the histograms represent the population parameters and the mean of the estimates, respectively. 




```{r message=FALSE, warning=F}

fit_all = lm(`right handspan (cm)`~`height (inches)`, data = dat)

results = data.frame(b0 = numeric(10000), b1 = numeric(10000))

for(i in 1:nrow(results)){
  dat_sample_c = sample_n(dat, 30)
  fit = lm(`right handspan (cm)`~`height (inches)`, data = dat_sample_c)
  results[i,1] = fit$coefficients[1]
  results[i,2] = fit$coefficients[2]
}

results |> ggplot(aes(x=b0))+
  geom_histogram() +
  geom_vline(xintercept=fit_all$coefficients[1], color = "blue")+
  geom_vline(xintercept = mean(results[,1]), color = "red")



```


```{r message=FALSE, warning=F}

results |> ggplot(aes(x=b1))+
  geom_histogram() +
  geom_vline(xintercept=fit_all$coefficients[2], color = "blue")+
  geom_vline(xintercept = mean(results[,2]), color = "red")

```



The equation of the population line is 
$$
y_i=`r fit_all$coefficients[1]`+`r fit_all$coefficients[2]` x_i
$$

The equation for the line for the first sample is 
$$
\hat{y}_i=`r fit_sample$coefficients[1]`+`r fit_sample$coefficients[2]` x_i
$$

Examining the histograms, we see the mean of all 10,000 $b_0$'s is `r mean(results[,1])` and the mean of all 10,000 $b_1$'s is `r mean(results[,2])`. 

We can get a good estimate of the sampling distributions by examining just  ten thousand samples. The downside to using repeated sampling is that we usually only have one sample. Thus, we need a different approach that will allow us to estimate the sampling distributions with just the information in our one sample. We will explore one way to do this next. 



### Bootstrap

*Bootstrapping* is a method for estimating the sampling distribution of a statistic based on the observations of *one sample*.

This estimation is done by sampling *with replacement* of size $n$ from the observed data. For each of these "bootstrap" samples, the estimator, both $\hat{\beta}_0$ and $\hat{\beta}_1$ in this case, is computed. This is done many times (usually thousands) so that the resulting distribution of these bootstrapped statistics provides an estimate for the sampling distribution of the statistic.


Suppose we only had a sample of size $n=30$ from the handspan and height data. This sample is plotted below with the least squares line (both in red).



```{r message=FALSE, warning=F}

boot_sample = sample_n(dat_sample, 30, replace=TRUE)

p2 = dat_sample |> ggplot(aes(x=`height (inches)`, y=`right handspan (cm)`)) +
  geom_point(color = "red") +
  geom_smooth(method="lm", se=FALSE, color = "red")

p2

```




A sample of size $n=30$ is taken from the red dots, with replacement. This bootstrapped sample are the black dots in the plot below. Note that some of the observations can be selected more than once due to sampling with replacement. This is why some of the points are still red.


```{r message=FALSE, warning=F}

boot_sample = sample_n(dat_sample, 30, replace=TRUE)

p2 + geom_point(data=boot_sample, color="black")+
  geom_smooth(data=boot_sample, method="lm", se=FALSE, color="black")


```



The equation of the line for the sample (the red line in the plot above) is
$$
\hat{y}_i=`r fit_sample$coefficients[1]`+`r fit_sample$coefficients[2]` x_i
$$


```{r message=FALSE, warning=F, echo=F}

fit_boot = lm(`right handspan (cm)`~`height (inches)`, data = boot_sample)

```

The equation for the line for the first bootstrap sample (the black line in the figure above) is
$$
\hat{y}_i=`r fit_boot$coefficients[1]`+`r fit_boot$coefficients[2]` x_i
$$


The $b_0$'s and $b_1$'s for 10,000 bootstrap samples are plotted in the histograms in the figure below.

```{r message=FALSE, warning=F}

results = data.frame(b0 = numeric(10000), b1 = numeric(10000))

for(i in 1:nrow(results)){
  boot_sample = sample_n(dat_sample, 30, replace=TRUE)
  fit = lm(`right handspan (cm)`~`height (inches)`, data = boot_sample)
  results[i,1] = fit$coefficients[1]
  results[i,2] = fit$coefficients[2]
}


results |> ggplot(aes(x=b0))+
  geom_histogram() +
  geom_vline(xintercept=fit_all$coefficients[1], color = "blue")+
  geom_vline(xintercept = fit_sample$coefficients[1], color = "red")



```


```{r message=FALSE, warning=F}

results |> ggplot(aes(x=b1))+
  geom_histogram() +
  geom_vline(xintercept=fit_all$coefficients[2], color = "blue")+
  geom_vline(xintercept = fit_sample$coefficients[2], color = "red")

```



Examining the histograms, we see the mean of all 10,000 $b_0$'s is `r mean(results[,1])` and the mean of all 10,000 $b_1$'s is `r mean(results[,2])`.

After generating 10,000 bootstrap estimates, we can get a good estimate of the sampling distributions of $b_0$ and $b_1$. Note, however, how these estimated distributions are centered at the least squares estimates of the observed sample (the red vertical lines) and not at the true population values (the blue vertical lines).


### Bootstrap Intervals


We just discussed how to  bootstrap the observations in order to estimated the sampling distributions for $b_0$ and $b_1$.

Based on the normality assumption, we could determine the sampling distributions for the coefficients theoretically without need to bootstrap.




When the normality assumption does not hold, then we can use the bootstrap to estimate the sampling distributions of the coefficients and for the mean response. We could also use it for the distribution of the predicted values.



Thus, we can still obtain confidence intervals for the coefficients, confidence intervals for the mean response, and prediction intervals even when normality does not hold.


#### Fixed Sampling vs Random Sampling


How the bootstrap works in regression is determined by the assumptions that hold for our model.



If the model is a good model for our data, the variance $\sigma^2$ is constant, and the predictor variables are regarded as fixed, then we use *fixed $x$* sampling.





In fixed $x$ sampling, the regression is fitted on the sample data and the fitted values, $\hat{y}_i$,  and the residuals, $e_i$ are obtained. A bootstrap sample of size $n$ are then obtained from the residuals which are denoted $e^*_i$. The bootstrap $y$ values are found as
$$
\begin{align}
y^*_i = \hat{y}_i +e^*_i
\end{align}
$${#eq-w5_17}


These bootstrapped $y_i^*$ are regressed on the original predictor variables. This procedure is done a large number of times and the resulting distributions of the coefficients $\hat{\beta}^*_i$ and fitted values $\hat{y}^*_i$ can be used as estimates of the corresponding sampling distributions.




If there is doubt in adequacy of the model, the variance $\sigma^2$ is not constant, and/or the predictor variables cannot be regarded as fixed, then we use *random $x$* sampling.



In random $x$ sampling, the observations (including $y$ and the predictor variables) are bootstrapped and then the resulting bootstrapped sample is used to fit the model.



::: {#exm-21_2}

## `fitness` data


We will examine the `fitness` dataset from the `olsrr` library.



In this dataset, we want to predict the oxygen level of the participants based on six predictor variables.


```{r message=FALSE, warning=F}
library(tidyverse)
library(tidymodels)
library(olsrr)

dat = fitness

dat_recipe = recipe(oxygen~., data = dat)

model = linear_reg() |> 
  set_engine("lm")

wf = workflow() |> 
  add_recipe(dat_recipe) |> 
  add_model(model)


# Create bootstrap samples
bootstraps = bootstraps(dat, times = 10)

# Fit the model to each bootstrap sample
boot_results <- bootstraps %>%
  mutate(
    model = map(splits, ~ fit(wf, data = analysis(.x))),
    coeff = map(model, ~ tidy(extract_fit_parsnip(.x))),
    mean_pred = map(splits, ~ predict(fit(wf, data = analysis(.x)), assessment(.x))),
    pred_interval = map(splits, ~ predict(fit(wf, data = analysis(.x)), assessment(.x), interval = "prediction"))
  )



# Extract and calculate percentile CIs for coefficients
coefficients <- bind_rows(boot_results$coeff)
ci_coefficients <- coefficients %>%
  group_by(term) %>%
  summarize(
    lower = quantile(estimate, 0.025),
    upper = quantile(estimate, 0.975)
  )

print(ci_coefficients)


```


:::




### Weighted Least Squares


Sometimes a transformation on $y$, such as a log transformation, can help stabilize nonconstant variance. However, if the linearity assumption between $y$ and the predictor variables seem reasonable, then transforming $y$ may violate the linearity assumption.



If we cannot transform $y$, we can adjust model @eq-w4_18
so that the variance term is allow to vary for different observations. Thus, we will have
$$
\begin{align}
{\textbf{Y}}= & {\textbf{X}}{\boldsymbol{\beta}}+{\boldsymbol{\varepsilon}}\\
& \boldsymbol{\varepsilon} \overset{iid}{\sim} N\left(0,\sigma_i^{2}\right)
\end{align}
$${#eq-w5_18}

The least squares estimators @eq-w4_21 could still be used. These estimators are *unbiased* but they
no longer have *minimum variance*. That is, they are no longer the BLUEs.






To obtain unbiased estimators with minimum variance,
we must take into account the different variances for the different $y$ observations. Observations with small variances provide more reliable information
about the regression function than those with large variances.





Therefore, we will want to *weight* the observations on the fit based on the variances of the observations.


### WLS Estimators


If the variances $\sigma_{i}^{2}$ are known, we can specify the *weights* as
$$
\begin{align}
w_{i} & =\frac{1}{\sigma_{i}^{2}}
\end{align}
$${#eq-w5_19}

We can specify diagonal weight matrix as
$$
\begin{align}
{\bf W} & =\left[\begin{array}{cccc}
w_{1} & 0 & \cdots & 0\\
0 & w_{2} & \cdots & 0\\
\vdots & \vdots & \ddots & \vdots\\
0 & 0 & \cdots & w_{n}
\end{array}\right]
\end{align}
$${#eq-w5_20}

These weights can be included in the least squares estimators as
$$
\begin{align}
{\bf b}_{w} & =\left({\bf X}^{\prime}{\bf W}{\bf X}\right)^{-1}{\bf X}^{\prime}{\bf W}{\bf Y}
\end{align}
$${#eq-w5_21}

The estimated covariance matrix of the *weighted* least squares (wls)
estimators is
$$
\begin{align}
{\bf s}^{2}\left[{\bf b}_{w}\right] & =\left({\bf X}^{\prime}{\bf W}{\bf X}\right)^{-1}
\end{align}
$${#eq-w5_22}

Compare this to the estimated covariance matrix for the *ordinary* (un-weighted)
least squares (ols) estimators in @eq-w4_31. Instead of using MSE for
an estimate of $\sigma^{2}$, @eq-w5_22 uses the weight matrix which
takes into account how the variance changes for different observations.



### The Variance and Standard Deviation Functions

For the wls estimators in @eq-w5_22, it is fairly straight forward if the variances $\sigma^2_i$ are known. In practice, these variances will be unknown and need to estimated.



Most of the time, the variance will vary in some systematic pattern. For example, the cone shape.




We can estimate a pattern like this by first fitting the model without any weights. We then examine the absolute values of the residuals vs the fitted values. We can regress these absolute residuals on the fitted values. This fitted model will be an estimate of the standard deviation function. Squaring these fitted values gives us an estimate of the variance function. The reciprocal of this variance function provides an estimate of the weights.




Other systematic patterns in the residual plots may require other ways to obtain an estimated variance function.




For example, if the plot of the residuals against $x_2$ suggests that the variance increases rapidly with
increases in $x_2$ up to a point and then increases more slowly, then we should  regress the absolute
residuals against $x_2$ and $x_2^2$.





If the wls estimates differs greatly from the ols estimates, it is common to take the squared or absolute residuals from the wls fit and then re-estimate the variance function. This is done again until the changes in the estimates become small between iterations.





This is known as *iteratively* *reweighted* least squares (irls).



::: {#exm-21_2}

## `bloodpressure` data

Let's revisit the blood pressure data used in @exm-w3_3_1. 

Let's first fit a regression model using typical ordinary least squares. 

```{r message=FALSE, warning=F}


dat = read.table("bloodpressure.txt", header=T)

dat_recipe = recipe(dbp~age, data = dat)

model = linear_reg() |> 
  set_engine("lm")

wf = workflow() |> 
  add_recipe(dat_recipe) |> 
  add_model(model)

fit_ols = wf |> fit(dat)

fit_ols |> tidy()

fit_ols |> extract_fit_engine() |>  ols_plot_resid_fit()
```

We see clear evidence of heteroscedasticity. Let's now use weighted least squares. We must first determine the weights. The absolute value of the residuals of the OLS fit is regressed on the fitted values of the OLS fit. The fitted values of this fit are then used to determine the weights.  



```{r message=FALSE, warning=F}

#fit absolute residuals against fitted values
results = fit_ols |> extract_fit_engine() |> augment()
fit_res = lm(abs(.resid)~.fitted, data = results)

#the fitted values will be used for the weights
w = 1/(fitted.values(fit_res)^2)

```


We can now add these weights to the workflow. We first specify that these are weights by adding it to the dataframe using `importance_weights`. 

```{r message=FALSE, warning=F}

dat = dat |> 
  mutate(
    weight = importance_weights(w)
  )

dat_recipe = recipe(dbp~age+weight, data = dat) 

wf = workflow() |> 
  add_case_weights(weight) |> 
  add_recipe(dat_recipe) |> 
  add_model(model)


fit_wls = wf |> fit(dat)

fit_wls |> tidy()

fit_ols |> tidy()

fit_wls |> glance()

fit_ols |> glance()

```

Note the difference between the estimates for the weighted least squares and the estimates for the ordinary least squares. The $R^2$ value is also higher using weighted least squares. 


:::

