

# Logistic Regression in Tidymodels

> "In God we trust, all others bring data." - W Edwards Deming



## Multiple Logistic Regression

In the last chapter, we introduced logistic regression using a simple (one predictor) model. The simple logistic regression model  is easily extended to more than one predictor
variable.





In fact, several predictor variables are usually required with logistic regression to
obtain adequate description and useful predictions.


In matrix notation, the logistic response function becomes
$$
E\{Y\} = \frac{\exp{\left(\textbf{X}^\prime\boldsymbol{\beta}\right)}}{1+\exp{\left(\textbf{X}^\prime\boldsymbol{\beta}\right)}}
$$


Like the simple logistic response function, the multiple logistic-response function
is monotonic and sigmoidal in shape with respect to $\textbf{X}^\prime\boldsymbol{\beta}$ and is almost linear
when $\pi$ is between .2 and .8.



The $X$ variables may be different predictor variables, or
some may represent curvature and/or interaction effects.


Also, the predictor variables may
be quantitative, or they may be qualitative and represented by indicator variables.


This
flexibility makes the multiple logistic regression model very useful.



## Example Using Tidymodels

In a health study to investigate an epidemic outbreak of a disease that is spread by mosquitoes, individuals were randomly sampled within two sectors in a city to determine if the person had recently contracted the disease under study.


This was ascertained by the interviewer,
who asked pertinent questions to assess whether certain specific symptoms associated with
the disease were present during the specified period.



The response variable $y$ was coded 1
if this disease was determined to have been present, and 0 if not.





Three predictor variables were included in the study, representing known or potential
risk factors.

They are age, socioeconomic status of household, and sector within city.

Age
($x_1$) is a quantitative variable. Socioeconomic status is a categorical variable with three
levels. It is represented by two indicator variables ($x_2$ and $x_3$), as follows:
$$
\begin{align*}
Class &\quad x_2 & x_3\\
Upper & \quad 0 & 0\\
Middle & \quad 1 & 0\\
Lower & \quad 0 & 1
\end{align*}
$$


City sector is also a categorical variable. Since there were only two sectors in the study,
one indicator variable ($x_4$) was used, defined so that $x_4 = 0$ for sector 1 and $x_4 = 1$ for
sector 2.

We will first convert the response variable to a factor and then look at the scatterplot matrix. 

```{r message=FALSE, warning=F}
library(tidyverse)
library(tidymodels)
library(GGally)

dat = read.table("CH14TA03.txt")

names(dat) = c("id", "X1", "X2", "X3", "X4", "Y")

dat = dat |> mutate(Y = relevel(as.factor(Y), ref="1"))

ggpairs(dat[,-1])

```


We can now setup the recipe, model, and workflow to fit the model.



```{r message=FALSE, warning=F}

dat_recipe = recipe(Y~X1+X2+X3+X4, data = dat) 

model = logistic_reg() |> 
  set_engine("glm")

wf = workflow() |> 
  add_recipe(dat_recipe) |> 
  add_model(model)

fit = wf |> fit(data=dat)

fit |> tidy()


```


The `tidy()` function returns the estimated values of the coefficients. Recall that we want to see the exponentials of the coefficients. 

```{r message=FALSE, warning=F}
fit |> tidy(exponentiate=T)

```

When we increase X1 (age) by one unit, we expect a 3% decrease in the odds of the disease being present, given the other predictors are held constant. 

For X2 being a 1 (middle class), we expect a 34% decrease in the odds of the disease being present, given the other predictors are held constant.

For X3 being a 1 (lower class), we expect a 36% increase in the odds of the disease being present, given the other predictors are held constant.

For X4 being a 1 (sector 2), we expect 79% decrease in the odds of the disease being present, given the other predictors are held constant. 

With the predict function, we can obtain the predicted probabilities of the fitted model.


## Predictions


```{r message=FALSE, warning=F}
pred_prob = predict(fit,
                    new_data = dat,
                    type = "prob")

pred_prob

```


If we classify each observation based on which in the most likely prediction, we can obtain the predicted outcomes.

```{r message=FALSE, warning=F}
pred_class = predict(fit,
                    new_data = dat,
                    type = "class")

pred_class

```

The value of the linear model can be obtained from the `predict` function. This value is what is obtained when the values of the predictor variables are substituted into the linear part of the model
$$
\hat{\beta}_0+\hat{\beta}_1 x_1 + \hat{\beta}_2 x_2 + \hat{\beta}_3 x_3 + \hat{\beta}_4 x_4
$$

```{r message=FALSE, warning=F}
pred_raw = predict(fit,
                    new_data = dat,
                    type = "raw")

pred_raw

```

These raw values can be used visualize the predicted probabilities and the actual values of the response variable.

```{r message=FALSE, warning=F}
results = dat |> 
  select(Y) |> 
  bind_cols(pred_raw, pred_prob, pred_class)

results |> 
  ggplot(aes(x = pred_raw, y = .pred_1, col = Y))+
  geom_point()+
  geom_hline(yintercept = 0.5)

```

A line at 0.5 represents the value for which we would classify the predicted probabilities. Values above 0.5 would be classified as a 1 (disease present) and values below 0.5 would be classified as a 0 (disease not present). The color of the dots are the actual values of the response variable. If the model was able to predict the response variable perfectly, then all of the values above 0.5 would be blue and all the values below 0.5 would be red. Since there are some red above the line and some blue below the line, then we have missclassification. 

## Confusion Matrix

To get a better idea of which is being missclassified, we can look at the confusion matrix.


```{r message=FALSE, warning=F}
conf_mat(results, truth = Y,
         estimate =  .pred_class)

```

We see from the confusion matrix that there are 9 observations that are predicted to have the disease but really did not (false positives). There were also 19 observations that were predicted to not have the disease but really did (false negatives). 

From this confusion matrix, we can calculate the probability of predicting a 1 given the observation is a 1 (a true positive). This is known as specificity or *True Positive Rate (TPR)*. 

Likewise, we can calculate the probability of prediction a 0 given the observation is a 0 (a true negative). This is known as specificity or *True Negative Rate (TNR)*. 

```{r message=FALSE, warning=F}
sens(results, truth = Y,
     estimate = .pred_class)

spec(results, truth = Y,
     estimate = .pred_class)

```




## ROC Curves

A Receiver Operating Characteristic (ROC) curve is a graphical representation of the diagnostic ability of a binary classifier. Logistic regression is such a classifier. 


The ROC curve plots the TPR versus the False Positive Rate (1-specificity). 

The ROC curve is created by calculating TPR and FPR at various threshold values, ranging from 0 to 1.

The *diagonal line* represents a random classifier (e.g., flipping a coin). A good classifier has a curve that bows significantly above this diagonal.

The *upper-left corner* of the plot (TPR = 1, FPR = 0) represents a perfect classifier, achieving both 100% sensitivity and 100% specificity.


#### Area Under the Curve (AUC)

The *Area Under the ROC Curve (AUC)* provides a single number summary of the model's performance:

AUC ranges from 0 to 1.

- **AUC = 1**: Perfect model.
- **AUC = 0.5**: Random guessing.
- **AUC < 0.5**: Worse than random guessing (usually indicates a problem with the model).

An ROC curve helps assess how well logistic regression separates the two classes, independent of the specific threshold chosen. It provides insights into the trade-offs between sensitivity and specificity as you adjust the threshold. Comparing ROC curves or AUC values of different models can guide model selection and threshold optimization.

To find the ROC curve:


```{r message=FALSE, warning=F}
results |> 
  roc_curve(truth = Y, .pred_1) |> 
  autoplot()

roc_auc(results, truth = Y, .pred_1)

```



















<!--  \subsection*{Multinomial Logistic Regression} -->


<!-- ## Nominal Response} -->

<!-- \marginpar{} -->
<!-- Logistic regression is most frequently used to model the relationship between a dichotomous -->
<!--                             response variable and a set of predictor variables. -->





<!--                             On occasion, however. the response -->
<!--                             variable may have *more} than two levels. -->





<!--                             Logistic regression can still be employed by -->
<!--                             means of a *polytomous}-or\\ *multinomial}-logistic regression model. -->





<!--                             Multinomial logistic -->
<!--                             regression models are used in many fields. In business, for instance, a market researcher -->
<!--                             may wish to relate a consumer's choice of product (product A, product B. product C) to -->
<!--                             the consumer's age, gender, geographic location and several other potential explanatory -->
<!--                             variables. -->





<!--                             This is an example of nominal multinomial regression because the response -->
<!--                             categories are purely qualitative and not *ordered} in any way. -->



<!-- *Ordinal} response categories can -->
<!--                             also be modeled using multinomial regression. -->





<!--                             For example, the relation between severity -->
<!--                             of disease measured on an ordinal scale (mild, moderate, severe) and age of patient, gender -->
<!--                             of patient, and some other explanatory variables may be of interest. -->





<!--                             The response variable $Y$ must be coded in the same way as indicator variables except we will need an indicator variable for each category of $Y$. -->





<!--                             Instead of going through the details, we will illustrate with R. -->




<!-- \begin{example} -->



<!-- A study was undertaken to determine the strength of association between several risk factors -->
<!--                             and the duration of pregnancies. -->



<!-- \marginpar{} -->

<!--                             The risk factors considered were mother's age, nutritional -->
<!--                             status, history of tobacco use, and history of alcohol use. -->





<!--                             The response of interest, pregnancy -->
<!--                             duration, is a three-category variable that was coded as follows: -->
<!--                             \begin{align*} -->
<!--                             y_{i} &\quad \textbf{Pregnancy Duration Category}\\ -->
<!--                             1 &\quad \text{Preterm (less than 36 weeks)}\\ -->
<!--                             2 &\quad \text{Intermediate term (36 to 37 weeks)}\\ -->
<!--                             3 &\quad \text{Full term (38 weeks or greater)} -->
<!--                             \end{align*} -->

<!--                             Relevant data for 102 women who had recently given birth at a large metropolitan hospital -->
<!--                             were obtained. -->





<!--                             X variables include: -->


<!--                             - Nutritional status ($x_1$), is an index of nutritional status (higher score denotes better nutritional status). -->

<!--                             - Age was categorized into three groups: less than 20 years of age (coded 1), from 21 -->
<!--                             to 30 years of age (coded 2), and greater than 30 years of age (coded 3). It is represented by -->
<!--                             two indicator variables (X2 and X3): -->
<!--                             \begin{align*} -->
<!--                             x_{2} &\quad x_{3} & \textbf{Class}\\ -->
<!--                             1 &\quad 0 & \text{Less than or equal to 20}\\ -->
<!--                             0 &\quad 0 & \text{21 to 30}\\ -->
<!--                             0 &\quad 1 & \text{Greater than 30} -->
<!--                             \end{align*} -->

<!--                             - Alcohol ($x_4$) and smoking history ($x_5$) were also qualitative predictors; the categories were "Yes" (coded 1) and "No" (coded 0). -->
<!-- \begin{lstlisting} -->
<!-- dat = read.table("http://users.stat.ufl.edu/~rrandles/sta4210/Rclassnotes/data/ -->
<!--       textdatasets/KutnerData/Chapter%2014%20Data%20Sets/CH14TA13.txt") -->
<!-- names(dat) = c("id", "Y", "Y1", "Y2", "Y3", "X1", "X2", "X3", "X4", "X5") -->

<!-- head(dat) -->

<!--   id Y Y1 Y2 Y3  X1 X2 X3 X4 X5 -->
<!-- 1  1 1  1  0  0 150  0  0  0  1 -->
<!-- 2  2 1  1  0  0 124  1  0  0  0 -->
<!-- 3  3 1  1  0  0 128  0  0  0  1 -->
<!-- 4  4 1  1  0  0 128  1  0  0  1 -->
<!-- 5  5 1  1  0  0 133  0  0  1  1 -->
<!-- 6  6 1  1  0  0 130  0  0  1  1 -->
<!-- \end{lstlisting} -->





<!--                             Note that the Y variable is already coded with the variables Y1, Y2, and Y3 in the dataset. -->

<!-- \marginpar{} -->
<!--                             We do not have to do this ourselves, R will code the indicator variables for us if we tell R the variable is a factor. -->



<!-- \begin{lstlisting} -->
<!-- library(nnet) -->

<!-- reg = multinom(factor(Y)~X1+X2+X3+X4+X5, data=dat) -->
<!-- summary(reg) -->

<!-- Call: -->
<!-- multinom(formula = factor(Y) ~ X1 + X2 + X3 + X4 + X5, data = dat) -->

<!-- Coefficients: -->
<!--   (Intercept)         X1          X2         X3         X4         X5 -->
<!-- 2   -1.517009 0.01897219 -0.04355881 -0.1721573 -0.9758857 -0.2218629 -->
<!-- 3   -5.475353 0.06542052 -2.95693397 -2.0596355 -2.0428964 -2.4522828 -->

<!-- Std. Errors: -->
<!--   (Intercept)         X1        X2        X3        X4        X5 -->
<!-- 2    2.047984 0.01631570 0.7373261 0.6943946 0.5869073 0.5856117 -->
<!-- 3    2.271691 0.01823932 0.9644844 0.8947673 0.7097457 0.7315046 -->

<!-- Residual Deviance: 168.6754 -->
<!-- AIC: 192.6754 -->
<!-- \end{lstlisting} -->






<!--                             Note in the summary, we have two regression fits. What we actually have done is model the odds of the probability of being in category 2 to the probability of being in category 1. -->





<!--                             We also modeled the odds of the probability of being in category 3 to the probability of being in category 1. -->





<!--                             So we call category 1 the baseline category. -->





<!--                             To get the effect on the odds ratio for each coefficient: -->



<!-- \begin{lstlisting}[firstnumber=last] -->
<!-- exp(coef(reg)) -->

<!--   (Intercept)       X1         X2        X3        X4         X5 -->
<!-- 2 0.219366942 1.019153 0.95737625 0.8418467 0.3768584 0.80102519 -->
<!-- 3 0.004188751 1.067608 0.05197804 0.1275004 0.1296526 0.08609682 -->
<!-- \end{lstlisting} -->




<!-- \marginpar{} -->

<!--                             We can use a different baseline category such as full term: -->



<!-- \begin{lstlisting}[firstnumber=last] -->
<!-- dat$Y = as.factor(dat$Y) -->
<!-- dat$Y = relevel(dat$Y, ref = "3") -->

<!-- reg.nom = multinom(factor(Y)~X1+X2+X3+X4+X5, data=dat) -->
<!-- summary(reg.nom) -->

<!-- Call: -->
<!-- multinom(formula = factor(Y) ~ X1 + X2 + X3 + X4 + X5, data = dat) -->

<!-- Coefficients: -->
<!--   (Intercept)          X1       X2       X3       X4       X5 -->
<!-- 1    5.475147 -0.06541919 2.957028 2.059662 2.042900 2.452362 -->
<!-- 2    3.958370 -0.04644903 2.913475 1.887550 1.067001 2.230492 -->

<!-- Std. Errors: -->
<!--   (Intercept)         X1        X2        X3        X4        X5 -->
<!-- 1    2.271677 0.01823916 0.9644921 0.8947727 0.7097461 0.7315106 -->
<!-- 2    1.941063 0.01488581 0.8575544 0.8088255 0.6495262 0.6681955 -->

<!-- Residual Deviance: 168.6754 -->
<!-- AIC: 192.6754 -->


<!-- #to get the predicted probability for each observation -->
<!-- preds.nom = fitted(reg.nom) -->
<!-- preds.nom[1:10,] -->

<!--             3           1          2 -->
<!-- 1  0.62078219 0.094217907 0.28499991 -->
<!-- 2  0.18455857 0.254210377 0.56123105 -->
<!-- 3  0.34297651 0.219535548 0.43748794 -->
<!-- 4  0.02716468 0.334553819 0.63828151 -->
<!-- 5  0.13335384 0.474690286 0.39195587 -->
<!-- 6  0.11480795 0.497290719 0.38790133 -->
<!-- 7  0.95144461 0.009569423 0.03898596 -->
<!-- 8  0.09738287 0.335195611 0.56742152 -->
<!-- 9  0.02775016 0.658559402 0.31369044 -->
<!-- 10 0.40441329 0.186642234 0.40894447 -->


<!-- # if we pedict the category with the highest probability: -->
<!-- ind = apply(preds.nom, 1, which.max) -->
<!-- preds.nom.cat = colnames(preds.nom)[ind] -->
<!-- preds.nom.cat -->

<!--   [1] "3" "2" "2" "2" "1" "1" "3" "2" "1" "2" "3" "1" "2" "1" "1" "3" "2" "2" "1" "1" "1" -->
<!--  [22] "1" "1" "2" "1" "2" "3" "3" "2" "2" "2" "2" "2" "2" "3" "2" "2" "2" "2" "3" "2" "2" -->
<!--  [43] "2" "2" "2" "2" "3" "2" "1" "2" "2" "1" "2" "3" "2" "2" "2" "1" "3" "1" "3" "3" "3" -->
<!--  [64] "3" "3" "3" "3" "3" "3" "3" "3" "3" "3" "3" "3" "3" "3" "3" "3" "3" "3" "3" "3" "3" -->
<!--  [85] "2" "3" "3" "3" "1" "3" "3" "3" "3" "3" "2" "2" "2" "2" "1" "1" "3" "1" -->



<!-- #compare to observed categories: -->
<!-- table(preds.nom.cat, dat$Y) -->

<!-- preds.nom.cat  3  1  2 -->
<!--             1  4 12  4 -->
<!--             2  5 10 23 -->
<!--             3 32  4  8 -->
<!-- \end{lstlisting} -->
<!-- \end{example} -->


<!--  \subsection*{Ordinal Regression} -->


<!-- ## Ordinal Response} -->


<!-- When the response variable $Y$ takes on categorical responses that can be put into order, then  the multinomial logistic regression discussed in Section 8.2 can still be used. -->
<!-- \marginpar{} -->




<!--                             However, a more effective strategy, yielding a more parsimonious and more easily interpreted model, results if the *ordering} of the categories is taken into account explicitly. -->





<!--                             The model that is usually employed is called the *proportional} *odds}  model -->





<!--                             The proportional odds model for ordinal logistic regression models the cumulative probabilities -->
<!--                             $$ -->
<!--                             P(y_i\le j) -->
<!--                             $$ -->
<!--                             rather than the specific category probabilities -->
<!--                             $$ -->
<!--                             P(y_i=j) -->
<!--                             $$ -->
<!--                             as was the case -->
<!--                             for nominal logistic regression. -->





<!--                             Again, we will skip the details and illustrate with R using the pregnancy data from the last section. -->
<!-- \marginpar{} -->
<!-- \begin{example} -->

<!-- \ -->


<!-- \begin{lstlisting} -->
<!-- library(MASS) -->

<!-- dat = read.table("http://users.stat.ufl.edu/~rrandles/sta4210/Rclassnotes/data/ -->
<!--                  textdatasets/KutnerData/Chapter%2014%20Data%20Sets/CH14TA13.txt") -->
<!-- names(dat) = c("id", "Y", "Y1", "Y2", "Y3", "X1", "X2", "X3", "X4", "X5") -->


<!-- #tell R the order of the factors -->
<!-- dat$Y = factor(dat$Y,levels = c("1", "2", "3"), ordered=T) -->

<!-- reg.ord = polr(Y~X1+X2+X3+X4+X5,data=dat) -->
<!-- summary(reg.ord) -->

<!-- Call: -->
<!-- polr(formula = Y ~ X1 + X2 + X3 + X4 + X5, data = dat) -->

<!-- Coefficients: -->
<!--       Value Std. Error t value -->
<!-- X1  0.04887    0.01182   4.133 -->
<!-- X2 -1.97601    0.57616  -3.430 -->
<!-- X3 -1.36348    0.54648  -2.495 -->
<!-- X4 -1.66987    0.47537  -3.513 -->
<!-- X5 -1.59154    0.45165  -3.524 -->

<!-- Intercepts: -->
<!--     Value   Std. Error t value -->
<!-- 1|2  2.9301  1.4929     1.9627 -->
<!-- 2|3  5.0249  1.5445     3.2535 -->

<!-- Residual Deviance: 173.5122 -->
<!-- AIC: 187.5122 -->


<!-- #to get the predicted probability for each observation -->
<!-- preds.ord = fitted(reg.ord) -->
<!-- preds.ord[1:10,] -->

<!--              1          2          3 -->
<!-- 1  0.056827016 0.27179270 0.67138028 -->
<!-- 2  0.239742367 0.47950086 0.28075677 -->
<!-- 3  0.150070798 0.43915298 0.41077622 -->
<!-- 4  0.560196430 0.35167985 0.08812372 -->
<!-- 5  0.423470637 0.43299765 0.14353172 -->
<!-- 6  0.459606967 0.41396070 0.12643233 -->
<!-- 7  0.009988634 0.06576689 0.92424447 -->
<!-- 8  0.384644784 0.45082830 0.16452692 -->
<!-- 9  0.751897195 0.20907082 0.03903198 -->
<!-- 10 0.121488681 0.40757743 0.47093389 -->


<!-- # if we pedict the category with the highest probability: -->
<!-- ind = apply(preds.ord, 1, which.max) -->
<!-- preds.ord.cat = colnames(preds.ord)[ind] -->
<!-- preds.ord.cat -->

<!--   [1] "3" "2" "2" "1" "2" "1" "3" "2" "1" "3" "3" "1" "2" "1" "1" "3" "2" "2" "1" "1" "1" -->
<!--  [22] "1" "1" "2" "1" "1" "3" "3" "1" "2" "2" "2" "2" "2" "3" "2" "2" "3" "2" "2" "2" "2" -->
<!--  [43] "1" "2" "2" "2" "3" "2" "2" "1" "2" "2" "1" "3" "1" "2" "1" "1" "2" "2" "3" "3" "3" -->
<!--  [64] "3" "3" "3" "3" "3" "3" "3" "2" "3" "3" "3" "3" "3" "3" "3" "3" "3" "3" "3" "3" "3" -->
<!--  [85] "3" "3" "3" "3" "2" "3" "2" "3" "3" "3" "3" "2" "2" "3" "2" "1" "3" "2" -->


<!-- #compare to observed categories: -->
<!-- table(preds.ord.cat, dat$Y) -->

<!-- preds.ord.cat  1  2  3 -->
<!--             1 13  7  1 -->
<!--             2  8 21  7 -->
<!--             3  5  7 33 -->
<!-- \end{lstlisting} -->
<!-- \end{example} -->



<!--  \subsection*{Poisson Regression} -->


<!-- ## Introduction} -->

<!-- \marginpar{} -->
<!-- We consider now another nonlinear regression model where the response outcomes are -->
<!--                             discrete. -->





<!--                             Poisson regression is useful when the outcome is a count, with large-count outcomes -->
<!--                             being rare events. -->





<!--                             For instance, the number of times a household shops at a particular -->
<!--                             supermarket in a week is a count, with a large number of shopping trips to the store during -->
<!--                             the week being a rare event. A researcher may wish to study the relation between a family's -->
<!--                             number of shopping trips to the store during a particular week and the family's income, -->
<!--                             number of children, distance from the store, and some other explanatory variables. -->





<!--                             As another -->
<!--                             example, the relation between the number of hospitalizations of a member of a health -->
<!--                             maintenance organization during the past year and the member's age, income, and previous -->
<!--                             health status may be of interest. -->






<!--                             The Poisson distribution can be utilized for outcomes that are counts ($y_i = 0, 1,2, ...$ ), -->
<!--                             with a large count or frequency being a rare event. -->

<!-- ## The Poisson Distribution} -->


<!-- The Poisson probability distribution is -->
<!--                             $$ -->
<!--                             f(Y) = \frac{\mu^Y \exp(-\mu)}{Y!} -->
<!--                             $$ -->

<!--                             The mean and variance of a Poisson distribution are -->
<!--                             \begin{align*} -->
<!--                             E\{Y\} &= \mu\\ -->
<!--                             \sigma^2\{Y\} &=\mu -->
<!--                             \end{align*} -->


<!--                             Note that the variance is the same as the mean. -->


<!-- \marginpar{} -->


<!--                             Hence, if the number of store trips follows -->
<!--                             the Poisson distribution and the mean number of store trips for a family with three children -->
<!--                             is larger than the mean number of trips for a family with no children, the variances of the -->
<!--                             distributions of outcomes for the two families will also differ. -->

<!-- ## Poisson Regression} -->


<!-- We start with the regression model -->
<!--                             \begin{align*} -->
<!--                             y_{i} & =E\left[ y_{i}\right] +\varepsilon_{i}\qquad i=1,2,\ldots,n -->
<!--                             \end{align*} -->


<!--                             The mean response for the $i$th case, to be denoted now my $\mu_{i}$ -->
<!--                             for simplicity, is assumed as always to be a function of the set of -->
<!--                             predictor variables $x_{1},\ldots,x_{p-1}$. -->





<!--                             We use the notation $\mu\left(\textbf{X}_{i},\boldsymbol{\beta}\right)$ -->
<!--                             to denote the function that relates the mean response $\mu_{i}$ to -->
<!--                             $\textbf{X}_{i}$, the values of the predictor variables for case -->
<!--                             $i$, and $\boldsymbol{\beta}$, the values of the regression coefficients. -->





<!--                             Some commonly used functions for Poisson regression are -->
<!--                             \begin{align*} -->
<!--                             \mu_{i}= & \mu\left(\textbf{X}_{i},\boldsymbol{\beta}\right)=\textbf{X}_{i}^{\prime}\boldsymbol{\beta}\\ -->
<!--                             \mu_{i}= & \mu\left(\textbf{X}_{i},\boldsymbol{\beta}\right)=\exp\left(\textbf{X}_{i}^{\prime}\boldsymbol{\beta}\right)\\ -->
<!--                             \mu_{i}= & \mu\left(\textbf{X}_{i},\boldsymbol{\beta}\right)=\ln\left(\textbf{X}_{i}^{\prime}\boldsymbol{\beta}\right) -->
<!--                             \end{align*} -->

<!--                             In all three cases, the mean response $\mu_{i}$ must be nonnegative. -->





<!--                             Since the distribution of the error terms $\varepsilon_{i}$ for Poisson -->
<!--                             regression is a function of the distribution of the response $y_{i}$, -->
<!--                             which is Poisson, it is easiest to state the Poisson regression model -->
<!--                             in the following form: -->

<!--                             $y_{i}$ are independent Poisson random variables with expected values $\mu_{i}$ where -->
<!--                             \begin{align*} -->
<!--                             \mu_{i} & =\mu\left(\textbf{X}_{i},\boldsymbol{\beta}\right) -->
<!--                             \end{align*} -->


<!--                             The most commonly used response function is $\mu_{i}=\exp\left(\textbf{X}_{i}^{\prime}\boldsymbol{\beta}\right)$. -->

<!-- \begin{example} -->

<!-- The Miller Lumber Company is a large retailer of lumber and paint, as well as of plumbing, -->
<!--                             electrical, and other household supplies. -->



<!-- \marginpar{} -->

<!--                             During a representative two-week period, in-store -->
<!--                             surveys were conducted and addresses of customers were obtained. The addresses were -->
<!--                             then used to identify the metropolitan area census tracts in which the customers reside. -->





<!--                             At -->
<!--                             the end of the survey period, the total number of customers who visited the store from each -->
<!--                             census tract within a 10-mile radius was determined and relevant demographic information -->
<!--                             for each tract (average income, number of housing units, etc.) was obtained. -->





<!--                             Several other -->
<!--                             variables expected to be related to customer counts were constructed from maps, including -->
<!--                             distance from census tract to nearest competitor and distance to store. -->






<!--                             Initial screening of the potential predictor variables was conducted which led to the -->
<!--                             retention of five predictor variables: -->


<!--                             - $x_1$: Number of housing units -->

<!--                             - $x_2$: Average income, in dollars -->

<!--                             - $x_3$: Average housing unit age, in years -->

<!--                             - $x_4$ : Distance to nearest competitor, in miles -->

<!--                             - $x_5$: Distance to store, in miles -->

<!--                             - $y_i$ : Number of customers who visited store from census tract -->


<!-- \begin{lstlisting} -->
<!-- dat = read.table("http://users.stat.ufl.edu/~rrandles/sta4210/Rclassnotes/data/ -->
<!--                textdatasets/KutnerData/Chapter%2014%20Data%20Sets/CH14TA14.txt") -->
<!-- names(dat) = c("Y", "X1", "X2", "X3", "X4", "X5") -->

<!-- reg = glm(Y~., family = "poisson", data=dat) -->
<!-- summary(reg) -->

<!-- Call: -->
<!-- glm(formula = Y ~ ., family = "poisson", data = dat) -->

<!-- Deviance Residuals: -->
<!--      Min        1Q    Median        3Q       Max -->
<!-- -2.93195  -0.58868  -0.00009   0.59269   2.23441 -->

<!-- Coefficients: -->
<!--               Estimate Std. Error z value Pr(>|z|) -->
<!-- (Intercept)  2.942e+00  2.072e-01  14.198  < 2e-16 *** -->
<!-- X1           6.058e-04  1.421e-04   4.262 2.02e-05 *** -->
<!-- X2          -1.169e-05  2.112e-06  -5.534 3.13e-08 *** -->
<!-- X3          -3.726e-03  1.782e-03  -2.091   0.0365 * -->
<!-- X4           1.684e-01  2.577e-02   6.534 6.39e-11 *** -->
<!-- X5          -1.288e-01  1.620e-02  -7.948 1.89e-15 *** -->
<!-- --- -->
<!-- Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1 -->

<!-- (Dispersion parameter for poisson family taken to be 1) -->

<!--     Null deviance: 422.22  on 109  degrees of freedom -->
<!-- Residual deviance: 114.99  on 104  degrees of freedom -->
<!-- AIC: 571.02 -->

<!-- Number of Fisher Scoring iterations: 4 -->

<!-- predict(reg) -->

<!--        1        2        3        4        5        6        7 -->
<!-- 2.512666 2.171006 3.336689 2.129078 1.982460 2.184004 1.458190 -->
<!--        8        9       10 -->
<!-- 2.397791 2.670277 2.453965 -->
<!-- \end{lstlisting} -->

<!-- \end{example} -->
<!-- \marginpar{} -->
